### 2018-07-02 paper.tex
stubs

\documentclass{article}
% Document
\begin{document}

\title{Infinite Cellular Automata Running Virtual 6502 Machines}
\author{Ian Holmes \\ Department of Bioengineering, University of California, Berkeley, USA}

\maketitle

\begin{abstract}
  Cellular automata.
  6502s.
  Blockchain.
\end{abstract}

\section{Introduction}

\cite{Nakamoto2008}

\section*{Acknowledgments}
Thanks to Chris Evans, Richard Evans, Michael Mateas.

\bibliographystyle{plain}
\bibliography{references}

\end{document}

### 2018-07-02 paper.tex
cube

@@
 \documentclass{article}
+
+% Macro to generate neighborhood memory map table
+\newcommand\memtable[1]{
+\begin{tabular}{lll}
+  \hline
+  From & To & Contents \\
+  \hline
+  #1
+  \hline
+\end{tabular}
+}
+
+\newcommand\memrow[4]{
+    {\tt {#1}00} & {\tt {#1}FF} & State of cell $(x#2,y#3,z#4)$ \\
+}
+
+% The \memxyz macro gives a densely-connected (von Neumann) neighborhood
+\newcommand\memz[5]{
+  \memrow{#1}{#4}{#5}{-1}
+  \memrow{#2}{#4}{#5}{}
+  \memrow{#3}{#4}{#5}{+1}
+}
+
+\newcommand\memyz[2]{
+  \memz{{#1}0}{{#1}1}{{#1}2}{#2}{-1}
+  \memz{{#1}4}{{#1}5}{{#1}6}{#2}{}
+  \memz{{#1}8}{{#1}9}{{#1}A}{#2}{+1}
+}
+
+\newcommand\memxyz{
+  \memyz{4}{-1}
+
+  %  \memyz{5}{} includes zero page-mapped cell (x,y,z), which we want to skip, so spell it out without that
+  \memz{50}{51}{52}{}{-1}
+  \memrow{54}{}{}{-1}
+  \memrow{56}{}{}{+1}
+  \memz{58}{59}{5A}{}{+1}
+
+  \memyz{6}{+1}
+}
+
+% The following table is a sparse neighborhood
+\newcommand\vonneumannmap{\memtable{
+  \memrow{45}{-1}{}{}
+  \memrow{51}{}{-1}{}
+  \memrow{54}{}{}{-1}
+  \memrow{56}{}{}{+1}
+  \memrow{59}{}{+1}{}
+  \memrow{65}{+1}{}{}
+}}
+
+% The following neighborhood is dense
+\newcommand\mooremap{\memtable{\memxyz}}
+
 % Document
@@
 
-\section{Introduction}
+\section{Cellular automata}
+
+The cellular automaton is a $L \times L \times L$ array of cells.
+Cells are indexed $(x,y,z$) where $0 \leq x \leq L$, $0 \leq y \leq L$, $0 \leq z \leq L$.
+Each cell has 256 bytes (one ``page'') of state.
+All cells' bytes are initially zero.
+
+Cells are updated in a pseudorandom order.
+This can be any order that visits all cells uniformly.
+If $L$ is a power of 2, $L = 2^N$ for some integer $N$,
+we can use the simple ordering $0 \leq i \leq 2^{3N}$
+with $(x_i,y_i,z_i)$ defined such that
+the $k$'th bit of $x_i$ is the $3N-3k-1$'th bit of $i$,
+the $k$'th bit of $y_i$ is the $3N-3k-2$'th bit of $i$, and
+the $k$'th bit of $z_i$ is the $3N-3k-3$'th bit of $i$
+(using LSB-0 numbering).
+
+A cell $(x,y,z)$ is updated by emulating an 8-bit virtual machine.
+The CPU is the MOS Technology 6502 processor.
+The memory map is shown in \ref{tab:MemoryMap},
+with detail for the memory-mapping of adjacent cells in \ref{tab:MooreNeighborhood}.
+
+\begin{table}
+\begin{tabular}{llll}
+  \hline
+  From & To & Type & Contents \\
+  \hline
+  {\tt 0000} & {\tt 00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
+  {\tt 0100} & {\tt 01FF} & RAM (transient) & Stack, initialized to zero \\
+  {\tt 0200} & {\tt 03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
+  {\tt 0400} & {\tt 0401} & ROM (emulator-mapped) & Cycles left before termination \\
+  {\tt 0402} & {\tt 0402} & ROM (emulator-mapped) & Random byte generator \\
+  {\tt 0400} & {\tt 3FFF} & ROM & Reserved, zero \\
+  {\tt 4000} & {\tt 7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
+  {\tt 8000} & {\tt FEFF} & ROM & Read-only (operating system) \\
+  {\tt FF00} & {\tt FFFF} & ROM & Includes interrupt and reset vectors. All zero \\
+  \hline
+\end{tabular}
+\caption{
+  \label{tab:MemoryMap}
+  Memory map for the 6502 virtual machine of cell $(x,y,z)$.
+  Note that the BREAK vector {\tt FFFC-FFFD} is zero, so execution begins at {\tt 0000}.
+}
+\end{table}
+
+\begin{table}
+\mooremap
+\caption{
+  \label{tab:MooreNeighborhood}
+  Memory map for the 6502 virtual machine of cell $(x,y,z)$,
+  using the 26-cell three-dimensional Moore neighborhood.
+}
+\end{table}
+
+\begin{table}
+\vonneumannmap
+\caption{
+  \label{tab:VonNeumannNeighborhood}
+  Memory map for the 6502 virtual machine of cell $(x,y,z)$,
+  using the 6-cell three-dimensional von Neumann neighborhood.
+}
+\end{table}
+
+All other bytes are read-only and zero; writing to them has no effect.
+This can be understood as the cell $(x-1+i,y-1+j,z-1+k)$
+mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
+Thus, in a memory-mapped 16-bit address, the bits have the following meanings
+
+\begin{tabular}{ll}
+  \hline
+  Bit(s) & Meaning \\
+  \hline
+  14,15 & Must be {\tt 01} \\
+  12,13 & $i$ \\
+  10,11 & $j$ \\
+  8,9 & $k$ \\
+  0-7 & Address within page \\
+  \hline
+\end{tabular}
+
+Note the exception: cell $(x,y,z)$ is not mapped to {\tt 5500-55FF}
+because it is mapped to zero page {\tt 0000-00FF} instead.
+
+\section{Blockchain description}
 

### 2018-07-02 paper.tex
cube

@@
 Cells are indexed $(x,y,z$) where $0 \leq x \leq L$, $0 \leq y \leq L$, $0 \leq z \leq L$.
+
 Each cell has 256 bytes (one ``page'') of state.
 All cells' bytes are initially zero.
+This is the ``storage'' of the virtual machine.
 
@@
 The CPU is the MOS Technology 6502 processor.
-The memory map is shown in \ref{tab:MemoryMap},
-with detail for the memory-mapping of adjacent cells in \ref{tab:MooreNeighborhood}.
+The memory map is shown in Table~\ref{tab:MemoryMap}.
+In this memory map, the 27 pages of storage
+for the cell and its 26 neighbors
+in the $3 \times 3 \times 3$ unit cube centered on $(x,y,z)$
+are mapped directly to pages in the RAM of the virtual machine.
+
+The detailed description of the memory map for this $3 \times 3 \times 3$ neighborhood,
+which in cellular automata terminology is
+the {\em three-dimensional Moore neighborhood},
+is given in Table~\ref{tab:MooreNeighborhood}.
+
+The memory map can be understood as the cell $(x+i-1,y+j-1,z+k-1)$
+mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
+Thus, in a memory-mapped 16-bit address, the bits have the following meanings
+
+\begin{tabular}{ll}
+  \hline
+  Bit(s) & Meaning \\
+  \hline
+  14,15 & Must be {\tt 01} \\
+  12,13 & $i$ \\
+  10,11 & $j$ \\
+  8,9 & $k$ \\
+  0-7 & Address within page \\
+  \hline
+\end{tabular}
+
+Note the exception: cell $(x,y,z)$ is not mapped to {\tt 5500-55FF}
+because it is mapped to zero page {\tt 0000-00FF} instead.
 
@@
   {\tt 0402} & {\tt 0402} & ROM (emulator-mapped) & Random byte generator \\
-  {\tt 0400} & {\tt 3FFF} & ROM & Reserved, zero \\
+  {\tt 0403} & {\tt 3FFF} & Reserved & Read-only, zero \\
   {\tt 4000} & {\tt 7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
@@
   using the 26-cell three-dimensional Moore neighborhood.
+  All address ranges not defined in this table are read-only and zero; writing to them has no effect.
 }
@@
 
-All other bytes are read-only and zero; writing to them has no effect.
-This can be understood as the cell $(x-1+i,y-1+j,z-1+k)$
-mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
-Thus, in a memory-mapped 16-bit address, the bits have the following meanings
-
-\begin{tabular}{ll}
-  \hline
-  Bit(s) & Meaning \\
-  \hline
-  14,15 & Must be {\tt 01} \\
-  12,13 & $i$ \\
-  10,11 & $j$ \\
-  8,9 & $k$ \\
-  0-7 & Address within page \\
-  \hline
-\end{tabular}
-
-Note the exception: cell $(x,y,z)$ is not mapped to {\tt 5500-55FF}
-because it is mapped to zero page {\tt 0000-00FF} instead.
-
 \section{Blockchain description}

### 2018-07-02 paper.tex
CA

@@
 \documentclass{article}
+
+% Memory address
+\newcommand\hex[1]{{\tt #1}}
+\newcommand\hexrange[2]{\hex{#1}{\tt -}\hex{#2}}
 
@@
 
-\title{Infinite Cellular Automata Running Virtual 6502 Machines}
-\author{Ian Holmes \\ Department of Bioengineering, University of California, Berkeley, USA}
+\title{Infinite Cellular Automata Running Virtual 6502 Processors}
+\author{Ian Holmes \\ Berkeley, California, USA}
 
@@
 
-The cellular automaton is a $L \times L \times L$ array of cells.
-Cells are indexed $(x,y,z$) where $0 \leq x \leq L$, $0 \leq y \leq L$, $0 \leq z \leq L$.
+The cellular automaton (CA) is a $L \times L \times L$ array of cells.
+Cells are indexed $(x,y,z$) where
+$\alpha \leq x \leq \alpha+L$,
+$\beta \leq y \leq \beta+L$,
+$\gamma \leq z \leq \gamma+L$
+and $(\alpha,\beta,\gamma)$ are integers representing global coordinates of this CA.
 
 Each cell has 256 bytes (one ``page'') of state.
-All cells' bytes are initially zero.
-This is the ``storage'' of the virtual machine.
+All cells' bytes are initially zero, unless otherwise specified
+(including, implicitly, cells whose coordinates are outside the CA).
+These bytes are the ``storage'' of the virtual machine.
 
-Cells are updated in a pseudorandom order.
-This can be any order that visits all cells uniformly.
-If $L$ is a power of 2, $L = 2^N$ for some integer $N$,
-we can use the simple ordering $0 \leq i \leq 2^{3N}$
-with $(x_i,y_i,z_i)$ defined such that
-the $k$'th bit of $x_i$ is the $3N-3k-1$'th bit of $i$,
-the $k$'th bit of $y_i$ is the $3N-3k-2$'th bit of $i$, and
-the $k$'th bit of $z_i$ is the $3N-3k-3$'th bit of $i$
-(using LSB-0 numbering).
+Cells in the CA are updated in a pseudorandom order.
+This can be any order that visits all cells uniformly;
+a Mersenne Twister is suitable.
 
@@
 in the $3 \times 3 \times 3$ unit cube centered on $(x,y,z)$
-are mapped directly to pages in the RAM of the virtual machine.
+are mapped directly to bytes in the address range \hexrange{4000}{7FFF} of the virtual machine.
 
@@
 
-The memory map can be understood as the cell $(x+i-1,y+j-1,z+k-1)$
+The neighborhood memory map can be understood as the cell $(x+i-1,y+j-1,z+k-1)$
 mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
-Thus, in a memory-mapped 16-bit address, the bits have the following meanings
+Thus, in a memory-mapped 16-bit address, the bits have the meanings shown in Table~\ref{tab:AddressBits}.
+Note the exception: cell $(x,y,z)$ is not mapped to page 85 (\hexrange{5500}{55FF})
+because it is mapped to zero page (\hexrange{0000}{00FF}) instead.
 
-\begin{tabular}{ll}
-  \hline
-  Bit(s) & Meaning \\
-  \hline
-  14,15 & Must be {\tt 01} \\
-  12,13 & $i$ \\
-  10,11 & $j$ \\
-  8,9 & $k$ \\
-  0-7 & Address within page \\
-  \hline
-\end{tabular}
+At the beginning of an update,
+the registers {\tt A,X,Y} are initialized to zero,
+all status flags (register {\tt P}) are set to zero,
+the stack pointer (register {\tt SP}) is set to \hex{FF},
+the program counter {\tt PC} is set to \hex{0000},
+and execution begins at the first byte of zero page.
 
-Note the exception: cell $(x,y,z)$ is not mapped to {\tt 5500-55FF}
-because it is mapped to zero page {\tt 0000-00FF} instead.
+Execution is terminated
+when a software interrupt or undefined instruction is encountered,
+or by non-maskable interrupt after a predetermined, but (pseudo)random, number of cycles $C$.
+The LSB of $C$ is uniformly distributed;
+the MSB is distributed according to a geometric distribution with parameter $\frac{1}{2}$.
+Thus, with probability $\frac{1}{2}$, the value of $C$ lies in the range \hexrange{0}{FF};
+with probability $\frac{1}{4}$, it lies in the range \hexrange{100}{1FF};
+with probability $\frac{1}{8}$ it's in the range \hexrange{200}{2FF};
+with probability $\frac{1}{16}$ it's \hexrange{300}{3FF};
+and so on.
+(It does not overflow when the MSB reaches \hex{FF}, which happens after the heat death of the universe.)
+Throughout execution, the {\em remaining} value of this timer
+(i.e. the number of cycles from now until termination)
+is visible to the program
+as the read-only 16-bit value at memory address \hexrange{0400}{0401}.
+
+On termination, the transient address ranges in the memory map
+(including the stack)
+are discarded.
+The neighborhood memory map (\hexrange{4000}{7FFF}, Table~\ref{tab:MooreNeighborhood})
+is mapped back to the cellular automata storage.
 
@@
   \hline
-  {\tt 0000} & {\tt 00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
-  {\tt 0100} & {\tt 01FF} & RAM (transient) & Stack, initialized to zero \\
-  {\tt 0200} & {\tt 03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
-  {\tt 0400} & {\tt 0401} & ROM (emulator-mapped) & Cycles left before termination \\
-  {\tt 0402} & {\tt 0402} & ROM (emulator-mapped) & Random byte generator \\
-  {\tt 0403} & {\tt 3FFF} & Reserved & Read-only, zero \\
-  {\tt 4000} & {\tt 7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
-  {\tt 8000} & {\tt FEFF} & ROM & Read-only (operating system) \\
-  {\tt FF00} & {\tt FFFF} & ROM & Includes interrupt and reset vectors. All zero \\
+  \hex{0000} & \hex{00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
+  \hex{0100} & \hex{01FF} & RAM (transient) & Stack, initialized to zero \\
+  \hex{0200} & \hex{03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
+  \hex{0400} & \hex{0401} & ROM (emulator-mapped) & Cycles left before termination \\
+  \hex{0402} & \hex{3FFF} & Reserved & Read-only, zero \\
+  \hex{4000} & \hex{7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
+  \hex{8000} & \hex{FFFF} & ROM & Read-only (operating system) \\
   \hline
@@
   Memory map for the 6502 virtual machine of cell $(x,y,z)$.
-  Note that the BREAK vector {\tt FFFC-FFFD} is zero, so execution begins at {\tt 0000}.
 }
@@
 
+\begin{table}
+\begin{tabular}{ll}
+  \hline
+  Bit(s) & Meaning \\
+  \hline
+  14,15 & Must be \hex{01} \\
+  12,13 & $i$ \\
+  10,11 & $j$ \\
+  8,9 & $k$ \\
+  0-7 & Address within page \\
+  \hline
+\end{tabular}
+\caption{
+  Interpretation of an address in the memory-mapped range \hexrange{4000}{7FFF}
+  as pointing to a cell $(x+i-1,y+j-1,z+k-1)$.
+  See Table~\ref{tab:MooreNeighborhood} and Table~\ref{tab:VonNeumannNeighborhood}.
+  Note that $i,j,k, \in \{ 0,1,2 \}$.
+  \label{tab:AddressBits}
+}
+\end{table}
+
 \section{Blockchain description}
@@
 \section*{Acknowledgments}
-Thanks to Chris Evans, Richard Evans, Michael Mateas.
+Thanks to Peter Irvine, Chris Evans, Richard Evans, Michael Mateas.
 

### 2018-07-02 paper.tex
CA

@@
 This can be any order that visits all cells uniformly;
-a Mersenne Twister is suitable.
+a Mersenne Twister is suitable for this,
+and also for the generation of the random variable {\tt RUNTIME} used to constrain the update (see below).
 
@@
 the program counter {\tt PC} is set to \hex{0000},
+{\tt TIMELEFT} is initalized to {\tt STOPTIME} (see below),
 and execution begins at the first byte of zero page.
@@
 when a software interrupt or undefined instruction is encountered,
-or by non-maskable interrupt after a predetermined, but (pseudo)random, number of cycles $C$.
-The LSB of $C$ is uniformly distributed;
+or by non-maskable interrupt after a predetermined, but (pseudo)random, number of cycles {\tt STOPTIME}.
+{\tt STOPTIME} is a 16-bit integer.
+The LSB of {\tt STOPTIME} is uniformly distributed;
 the MSB is distributed according to a geometric distribution with parameter $\frac{1}{2}$.
-Thus, with probability $\frac{1}{2}$, the value of $C$ lies in the range \hexrange{0}{FF};
+Thus, with probability $\frac{1}{2}$, the value of {\tt STOPTIME} lies in the range \hexrange{0}{FF};
 with probability $\frac{1}{4}$, it lies in the range \hexrange{100}{1FF};
@@
 and so on.
-(It does not overflow when the MSB reaches \hex{FF}, which happens after the heat death of the universe.)
+({\tt STOPTIME} stops incrementing and does not overflow when the MSB reaches \hex{FF},
+which happens after the heat death of the universe.)
 Throughout execution, the {\em remaining} value of this timer
@@
 is visible to the program
-as the read-only 16-bit value at memory address \hexrange{0400}{0401}.
+as the read-only 16-bit value called {\tt TIMELEFT},
+available at memory address \hexrange{0400}{0401}.
 
@@
   \hex{0200} & \hex{03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
-  \hex{0400} & \hex{0401} & ROM (emulator-mapped) & Cycles left before termination \\
+  \hex{0400} & \hex{0401} & ROM (emulator-mapped) & {\tt TIMELEFT}, cycles left before termination \\
   \hex{0402} & \hex{3FFF} & Reserved & Read-only, zero \\

### 2018-07-02 paper.tex
CA

@@
 Cells are indexed $(x,y,z$) where
-$\alpha \leq x \leq \alpha+L$,
-$\beta \leq y \leq \beta+L$,
-$\gamma \leq z \leq \gamma+L$
-and $(\alpha,\beta,\gamma)$ are integers representing global coordinates of this CA.
+$\alpha_x \leq x < \alpha_x+L$,
+$\alpha_y \leq y < \alpha_y+L$,
+$\alpha_z \leq z < \alpha_z+L$
+and $(\alpha_x,\alpha_y,\alpha_z)$ are integers representing global coordinates of this CA.
 
@@
 
-Cells in the CA are updated in a pseudorandom order.
+An {\bf update} of the CA consists of updating each cell in the CA in a pseudorandom order.
 This can be any order that visits all cells uniformly;
@@
 The neighborhood memory map (\hexrange{4000}{7FFF}, Table~\ref{tab:MooreNeighborhood})
-is mapped back to the cellular automata storage.
+is copied back to the cellular automata storage.
+
+\section{Adjacencies}
+
+Consider those cell coordinates $(x,y,z)$ that are outside the CA,
+e.g. $(\alpha_x-1,\alpha_y,\alpha_z)$ or  $(\alpha_x,\alpha_y,\alpha_z + L)$.
+Some of these, specifically those in the Moore neighborhood of cells that are in the CA,
+are a relevant initial condition for the machine,
+since they map to bytes in its addressable memory.
 
@@
   All address ranges not defined in this table are read-only and zero; writing to them has no effect.
+  Pages mapping to cells outside the CA (e.g. $x<\alpha_x$ or $x \geq \alpha_x+L$, or similarly for $y,z$)
+  are zero when read (unless otherwise specified),
+  and writing to them has no effect.
 }
@@
   using the 6-cell three-dimensional von Neumann neighborhood.
+  This memory map may be considered a subset of the memory map in Table~\ref{tab:MooreNeighborhood}.
 }

### 2018-07-02 paper.tex
edits

@@
 
-\title{Infinite Cellular Automata Running Virtual 6502 Processors}
+\title{Peer-to-Peer 6502 Cellular Automata}
 \author{Ian Holmes \\ Berkeley, California, USA}
@@
 \begin{abstract}
-  Cellular automata.
-  6502s.
-  Blockchain.
+  As a toy illustration,
+  this paper describes a blockchain cellular automata
+  comprising a 3D array of virtual machines.
+  Each machine emulates a 6502 processor with access to one page (256 bytes) of persistent storage,
+  as well as the pages of its neighbors (and a read-only operating system).
+  Simulation of this machine is incentivized via proof-of-work,
+  writes are limited by transaction fee,
+  and communications between adjacent nodes on the 3D grid are modeled as blockchain mergers.
 \end{abstract}
@@
 
-The cellular automaton (CA) is a $L \times L \times L$ array of cells.
-Cells are indexed $(x,y,z$) where
+The cellular automaton (CA) is a $L \times L \times L$ array of cells,
+located at an offset $(\alpha_x,\alpha_y,\alpha_z)$ in integer 3-space.
+Cells in the CA are indexed $(x,y,z$) where
 $\alpha_x \leq x < \alpha_x+L$,
 $\alpha_y \leq y < \alpha_y+L$,
-$\alpha_z \leq z < \alpha_z+L$
-and $(\alpha_x,\alpha_y,\alpha_z)$ are integers representing global coordinates of this CA.
+$\alpha_z \leq z < \alpha_z+L$.
 
@@
 An {\bf update} of the CA consists of updating each cell in the CA in a pseudorandom order.
-This can be any order that visits all cells uniformly;
+This can be any order that visits $L^3$ uniformly-random selected cells;
 a Mersenne Twister is suitable for this,
@@
 the MSB is distributed according to a geometric distribution with parameter $\frac{1}{2}$.
-Thus, with probability $\frac{1}{2}$, the value of {\tt STOPTIME} lies in the range \hexrange{0}{FF};
+Thus, with probability $\frac{1}{2}$, the value of {\tt STOPTIME} lies uniformly in the range \hexrange{00}{FF};
 with probability $\frac{1}{4}$, it lies in the range \hexrange{100}{1FF};
@@
 
-\section{Adjacencies}
+\section{Blankets}
+\label{sec:Blankets}
 
@@
 
+We refer to these cells as the {\bf blanket} of the CA.
+Previously, we specified that their state was implicitly initialized to zero.
+More generally, we may refer to the initial (and perpetual, throughout the execution of an update)
+state of these cells as the CA's {\bf blanket state}.
+This blanket state is zero until otherwise specified, but may be affected
+by interactions with neighboring CAs in the infinite vector space,
+as for example by the blockchain of Section~\ref{sec:Blockchain}.
+
+\section{Blockchain}
+\label{sec:Blockchain}
+
+The CA has an associated blockchain, defined as in \cite{Nakamoto2008}.
+The proof-of-work is adjusted to include updating the CA,
+and the mining incentive is adjusted to include redemption of coins minted by adjacent CAs
+(this also updates the blanket cells accounted for by those CAs).
+
+Specifically, let the proof-of-work include $C \times Z$ updates of the CA,
+where $Z$ is the number of zeroes required by the network at the start of the hash function for a valid block,
+and $C$ is the number of cycles expected in the blockchain update time
+(which for bitcoin is 10 minutes).
+
+Let the definition of a block include the coordinates $(\alpha_x,\alpha_y,\alpha_z,L)$;
+these coordinates, together with the number of updates,
+must be used to seed the random number generator for the $Z$ updates,
+and the block must include the $256 L^3$ bytes describing the state of the CA following the update.
+
+Any transaction in the blockchain may include, as additional data,
+a {\bf write operation} which modifies the state of one or more cells of the CA.
+This modification is considered to take place after the emulation of $Z$ update cycles
+which is part of the proof-of-work protocol.
+The order in which such write operations are applied,
+each potentially overwriting the state modifications of its predecessors,
+is implied by the ordering of the transaction in the block.
+As usual, the transaction may include network fees to incentivize inclusion of transactions
+by the miner
+(and prioritization later in the block).
+The transaction fee is, as usual, the difference between the output and input values of the transaction,
+minus a write operation tax of $W$ coins per cell written,
+to discourage frivolous edits.
+(Suggest $W=1$ to keep this simple.
+Minting a coin lets you write a cell.
+Seems fair.)
+
+The block may include sections of blockchain from {\bf adjacent cellular automata},
+defined to be CAs with co-ordinates containing cells that
+overlap with the current CA's blanket,
+but no cells that overlap with
+cells in the co-ordinate range of this CA (or any other CA previously encountered in the blockchain,
+except the adjacent CA itself).
+These blockchain segments from adjacent CAs are called {\bf adjacent segments}.
+
+Adjacent segments must be consistent with any previously-observed adjacent segments
+earlier on the current blockchain (that is, they must not involve conflicting CA,
+and they must otherwise be consistent continuations of the adjacent CA's blockchain).
+They may, themselves, recursively refer to parts of the current blockchain as their own adjacent segments,
+summarized by their hashes for compactness.
+
+When a block includes an adjacent segment, the CA's blanket is updated to reflect the state
+of the blanket cells described by the adjacent segment.
+Regular inclusion of adjacent segments from an adjacent CA's blockchain
+corresponds to keeping a bidirectional read-only interface open with that CA.
+
+To incentivize this,
+inclusion of an adjacent segment in a block for the first time
+accrues an additional benefit of $C/R$ coins to the miner of the block,
+where $C$ is the number of {\bf redeemable coins} mined or accrued (via more distant, proximal adjacencies)
+by cells in the adjacent CA,
+and $R$ is the {\bf redemption rate}, the number of redeemable coins in adjacent CAs
+required to accrue a coin in the current CA.
+It is of course necessary to have $R>2$ to prevent an explosion of redeemable coins in 3D space;
+another notable value is $R=26$, which (in the densely-connected Moore neighborhood)
+ensures that the number of coins redeemed is less than the number of coins mined locally
+(the corresponding value for the von Neumann neighborhood of Table~\ref{tab:VonNeumannNeighborhood}
+is much lower, at $R=6$).
+
+Setting the parameters $L$ and $C$ is a little arbitrary.
+If we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of 400 cycles per cellular update
+(the random variable {\tt STOPTIME} actually has a mean of 383.5 cycles),
+we would update the whole board $10^6 \times 10 \times 60 \times 60 / (400 \times L^3)$ times
+using the 10-minute bitcoin block rate.
+The CA state requires $256 \times L^3$ bytes.
+Faster clocks and protocols no doubt allow better numbers,
+but considering these values and estimating conservatively,
+we recommend $L=64$ and $C=256$ so that each block takes at most 64Mb to describe the CA state;
+each cell is updated, on average, every few minutes;
+and the proof-of-work burden (for $Z=1$) is roughly comparable to a 6502 CPU.
+
+% Tables
 \begin{table}
@@
   \hex{0200} & \hex{03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
-  \hex{0400} & \hex{0401} & ROM (emulator-mapped) & {\tt TIMELEFT}, cycles left before termination \\
+  \hex{0400} & \hex{0401} & ROM (emulator-mapped) & {\tt TIMELEFT}: cycles left before termination \\
   \hex{0402} & \hex{3FFF} & Reserved & Read-only, zero \\
@@
   using the 26-cell three-dimensional Moore neighborhood.
-  All address ranges not defined in this table are read-only and zero; writing to them has no effect.
+  All address ranges not defined in this table are read-only and zero;
+  writing to these addresses has no effect.
   Pages mapping to cells outside the CA (e.g. $x<\alpha_x$ or $x \geq \alpha_x+L$, or similarly for $y,z$)
-  are zero when read (unless otherwise specified),
-  and writing to them has no effect.
+  are read-only and zero (unless otherwise specified, as in Section~\ref{sec:Blankets}).
+  Writing to these addresses has no effect.
 }
@@
 
-\section{Blockchain description}
-
-\cite{Nakamoto2008}
-
 \section*{Acknowledgments}

### 2018-07-03 paper.tex
tidying

@@
 
-\section{Cellular automata}
+\section{Cellular Automaton}
+
+\subsection{Storage}
+\label{sec:Storage}
 
@@
 
-Each cell has 256 bytes (one ``page'') of state.
+Each cell has 256 bytes (one page) of state.
 All cells' bytes are initially zero, unless otherwise specified
-(including, implicitly, cells whose coordinates are outside the CA).
-These bytes are the ``storage'' of the virtual machine.
+(including, implicitly, cells whose coordinates are outside the CA; see also Section~\ref{sec:Blankets}).
+These bytes are the storage of the virtual machine.
+
+\subsection{Virtual machine}
 
@@
 
-\section{Blankets}
+\subsection{Blankets}
 \label{sec:Blankets}
@@
 We refer to these cells as the {\bf blanket} of the CA.
-Previously, we specified that their state was implicitly initialized to zero.
+Earlier, in Section~\ref{sec:Storage}, we specified that the state of these cells was implicitly initialized to zero.
 More generally, we may refer to the initial (and perpetual, throughout the execution of an update)
@@
 The CA has an associated blockchain, defined as in \cite{Nakamoto2008}.
-The proof-of-work is adjusted to include updating the CA,
-and the mining incentive is adjusted to include redemption of coins minted by adjacent CAs
+Updates are incentivized by mining,
+and the reward for mining is adjusted to include redemption of coins mined by adjacent CA blockchains
 (this also updates the blanket cells accounted for by those CAs).
 
-Specifically, let the proof-of-work include $C \times Z$ updates of the CA,
-where $Z$ is the number of zeroes required by the network at the start of the hash function for a valid block,
-and $C$ is the number of cycles expected in the blockchain update time
+\subsection{Updates}
+
+A valid block must include the $256 \times L^3$ bytes describing the state of the CA following $C$ updates
+from the state at the end of the previous block.
+$C$ is the number of cycles per cell expected in the blockchain update time
 (which for bitcoin is 10 minutes).
@@
 these coordinates, together with the number of updates,
-must be used to seed the random number generator for the $Z$ updates,
-and the block must include the $256 L^3$ bytes describing the state of the CA following the update.
+must be used to seed the random number generator for the $Z$ updates.
+
+\subsection{Writes}
 
@@
 Seems fair.)
+
+\subsection{Adjacencies}
 

### 2018-07-03 paper.tex
removed quotes
CA

@@
 
-\section{Cellular Automaton}
+\section{Cellular automata}
 
@@
 All cells' bytes are initially zero, unless otherwise specified
-(including, implicitly, cells whose coordinates are outside the CA; see also Section~\ref{sec:Blankets}).
+(including, implicitly, cells whose coordinates are outside the CA; see also Section~\ref{sec:AdjacentCells}).
 These bytes are the storage of the virtual machine.
@@
 
-\subsection{Blankets}
-\label{sec:Blankets}
+\subsection{Adjacent cells}
+\label{sec:AdjacentCells}
 
@@
 
-We refer to these cells as the {\bf blanket} of the CA.
+We refer to these cells as the CA's {\bf adjacent cells}.
 Earlier, in Section~\ref{sec:Storage}, we specified that the state of these cells was implicitly initialized to zero.
 More generally, we may refer to the initial (and perpetual, throughout the execution of an update)
-state of these cells as the CA's {\bf blanket state}.
-This blanket state is zero until otherwise specified, but may be affected
+state of these cells as the CA's {\bf adjacent state}.
+This adjacent state is zero until otherwise specified, but may be affected
 by interactions with neighboring CAs in the infinite vector space,
@@
 and the reward for mining is adjusted to include redemption of coins mined by adjacent CA blockchains
-(this also updates the blanket cells accounted for by those CAs).
+(this also updates the adjacent cell states accounted for by those CAs).
 
@@
 
-\subsection{Writes}
+\subsection{Write operations}
 
@@
 
-\subsection{Adjacencies}
+\subsection{Adjacent segments}
 
 The block may include sections of blockchain from {\bf adjacent cellular automata},
-defined to be CAs with co-ordinates containing cells that
-overlap with the current CA's blanket,
+defined to be CAs with co-ordinate ranges that
+overlap with the current CA's adjacent cells,
 but no cells that overlap with
@@
 
-When a block includes an adjacent segment, the CA's blanket is updated to reflect the state
-of the blanket cells described by the adjacent segment.
-Regular inclusion of adjacent segments from an adjacent CA's blockchain
-corresponds to keeping a bidirectional read-only interface open with that CA.
+When a block includes an adjacent segment, the CA's adjacent cell state is updated to match the adjacent segment.
+Regular mutual inclusion of adjacent segments from adjacent CAs' blockchains
+corresponds to keeping a bidirectional read-only interface open between those CAs.
 
@@
   \hex{0100} & \hex{01FF} & RAM (transient) & Stack, initialized to zero \\
-  \hex{0200} & \hex{03FF} & RAM (transient) & ``Scratch'' memory, initialized to zero \\
+  \hex{0200} & \hex{03FF} & RAM (transient) & Scratch memory, initialized to zero \\
   \hex{0400} & \hex{0401} & ROM (emulator-mapped) & {\tt TIMELEFT}: cycles left before termination \\
@@
   Pages mapping to cells outside the CA (e.g. $x<\alpha_x$ or $x \geq \alpha_x+L$, or similarly for $y,z$)
-  are read-only and zero (unless otherwise specified, as in Section~\ref{sec:Blankets}).
+  are read-only and zero (unless otherwise specified, as in Section~\ref{sec:AdjacentCells}).
   Writing to these addresses has no effect.

### 2018-07-03 paper.tex
CA
abstract

@@
   As a toy illustration,
-  this paper describes a blockchain cellular automata
+  this paper describes a blockchain for cellular automata
   comprising a 3D array of virtual machines.
@@
   as well as the pages of its neighbors (and a read-only operating system).
-  Simulation of this machine is incentivized via proof-of-work,
+  Simulation of this machine is incentivized via mining,
   writes are limited by transaction fee,

### 2018-07-03 paper.tex
removed time

@@
 \newcommand\hexrange[2]{\hex{#1}{\tt -}\hex{#2}}
-
-% Macro to generate neighborhood memory map table
-\newcommand\memtable[1]{
-\begin{tabular}{lll}
-  \hline
-  From & To & Contents \\
-  \hline
-  #1
-  \hline
-\end{tabular}
-}
-
-\newcommand\memrow[4]{
-    {\tt {#1}00} & {\tt {#1}FF} & State of cell $(x#2,y#3,z#4)$ \\
-}
-
-% The \memxyz macro gives a densely-connected (von Neumann) neighborhood
-\newcommand\memz[5]{
-  \memrow{#1}{#4}{#5}{-1}
-  \memrow{#2}{#4}{#5}{}
-  \memrow{#3}{#4}{#5}{+1}
-}
-
-\newcommand\memyz[2]{
-  \memz{{#1}0}{{#1}1}{{#1}2}{#2}{-1}
-  \memz{{#1}4}{{#1}5}{{#1}6}{#2}{}
-  \memz{{#1}8}{{#1}9}{{#1}A}{#2}{+1}
-}
-
-\newcommand\memxyz{
-  \memyz{4}{-1}
-
-  %  \memyz{5}{} includes zero page-mapped cell (x,y,z), which we want to skip, so spell it out without that
-  \memz{50}{51}{52}{}{-1}
-  \memrow{54}{}{}{-1}
-  \memrow{56}{}{}{+1}
-  \memz{58}{59}{5A}{}{+1}
-
-  \memyz{6}{+1}
-}
-
-% The following table is a sparse neighborhood
-\newcommand\vonneumannmap{\memtable{
-  \memrow{45}{-1}{}{}
-  \memrow{51}{}{-1}{}
-  \memrow{54}{}{}{-1}
-  \memrow{56}{}{}{+1}
-  \memrow{59}{}{+1}{}
-  \memrow{65}{+1}{}{}
-}}
-
-% The following neighborhood is dense
-\newcommand\mooremap{\memtable{\memxyz}}
 
@@
 This can be any order that visits $L^3$ uniformly-random selected cells;
-a Mersenne Twister is suitable for this,
-and also for the generation of the random variable {\tt RUNTIME} used to constrain the update (see below).
+a Mersenne Twister is suitable for this.
 
@@
 the program counter {\tt PC} is set to \hex{0000},
-{\tt TIMELEFT} is initalized to {\tt STOPTIME} (see below),
 and execution begins at the first byte of zero page.
@@
 when a software interrupt or undefined instruction is encountered,
-or by non-maskable interrupt after a predetermined, but (pseudo)random, number of cycles {\tt STOPTIME}.
-{\tt STOPTIME} is a 16-bit integer.
-The LSB of {\tt STOPTIME} is uniformly distributed;
-the MSB is distributed according to a geometric distribution with parameter $\frac{1}{2}$.
-Thus, with probability $\frac{1}{2}$, the value of {\tt STOPTIME} lies uniformly in the range \hexrange{00}{FF};
-with probability $\frac{1}{4}$, it lies in the range \hexrange{100}{1FF};
-with probability $\frac{1}{8}$ it's in the range \hexrange{200}{2FF};
-with probability $\frac{1}{16}$ it's \hexrange{300}{3FF};
-and so on.
-({\tt STOPTIME} stops incrementing and does not overflow when the MSB reaches \hex{FF},
-which happens after the heat death of the universe.)
-Throughout execution, the {\em remaining} value of this timer
-(i.e. the number of cycles from now until termination)
-is visible to the program
-as the read-only 16-bit value called {\tt TIMELEFT},
-available at memory address \hexrange{0400}{0401}.
+or by non-maskable interrupt after
+completion of the first instruction that takes the cycle count past 65,536 cycles.
 
@@
 
-Setting the parameters $L$ and $C$ is a little arbitrary.
-If we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of 400 cycles per cellular update
-(the random variable {\tt STOPTIME} actually has a mean of 383.5 cycles),
-we would update the whole board $10^6 \times 10 \times 60 \times 60 / (400 \times L^3)$ times
-using the 10-minute bitcoin block rate.
-The CA state requires $256 \times L^3$ bytes.
+Setting the parameters $L$ (CA size) and $C$ (updates per block)
+is a little arbitrary, but can be guided by target memory footprint and clock rate.
+For memory, the CA state requires $256 \times L^3$ bytes.
+For clock rate, if we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of $2^{15}$ cycles per cellular update,
+we would update the whole board $10^6 \times T / (2^{15} \times L^3)$ times
+where $T$ is the time in seconds between blocks ($T=36,000$ for bitcoin).
 Faster clocks and protocols no doubt allow better numbers,
 but considering these values and estimating conservatively,
-we recommend $L=64$ and $C=256$ so that each block takes at most 64Mb to describe the CA state;
-each cell is updated, on average, every few minutes;
-and the proof-of-work burden (for $Z=1$) is roughly comparable to a 6502 CPU.
+we recommend $L=64$ and $C=4$ so that
+each cell is updated every few minutes;
+each block takes at most 64Mb to describe the CA state;
+and the computational burden is roughly comparable to a 6502 CPU.
+
+% Tables, macros
+% Macro to generate neighborhood memory map table
+\newcommand\memtable[1]{
+\begin{tabular}{lll}
+  \hline
+  From & To & Contents \\
+  \hline
+  #1
+  \hline
+\end{tabular}
+}
+
+\newcommand\memrow[4]{
+    {\tt {#1}00} & {\tt {#1}FF} & State of cell $(x#2,y#3,z#4)$ \\
+}
+
+% The \memxyz macro gives a densely-connected (von Neumann) neighborhood
+\newcommand\memz[5]{
+  \memrow{#1}{#4}{#5}{-1}
+  \memrow{#2}{#4}{#5}{}
+  \memrow{#3}{#4}{#5}{+1}
+}
+
+\newcommand\memyz[2]{
+  \memz{{#1}0}{{#1}1}{{#1}2}{#2}{-1}
+  \memz{{#1}4}{{#1}5}{{#1}6}{#2}{}
+  \memz{{#1}8}{{#1}9}{{#1}A}{#2}{+1}
+}
+
+\newcommand\memxyz{
+  \memyz{4}{-1}
+
+  %  \memyz{5}{} includes zero page-mapped cell (x,y,z), which we want to skip, so spell it out without that
+  \memz{50}{51}{52}{}{-1}
+  \memrow{54}{}{}{-1}
+  \memrow{56}{}{}{+1}
+  \memz{58}{59}{5A}{}{+1}
+
+  \memyz{6}{+1}
+}
+
+% The following table is a sparse neighborhood
+\newcommand\vonneumannmap{\memtable{
+  \memrow{45}{-1}{}{}
+  \memrow{51}{}{-1}{}
+  \memrow{54}{}{}{-1}
+  \memrow{56}{}{}{+1}
+  \memrow{59}{}{+1}{}
+  \memrow{65}{+1}{}{}
+}}
+
+% The following neighborhood is dense
+\newcommand\mooremap{\memtable{\memxyz}}
 
@@
   \hex{0200} & \hex{03FF} & RAM (transient) & Scratch memory, initialized to zero \\
-  \hex{0400} & \hex{0401} & ROM (emulator-mapped) & {\tt TIMELEFT}: cycles left before termination \\
-  \hex{0402} & \hex{3FFF} & Reserved & Read-only, zero \\
+  \hex{0400} & \hex{3FFF} & ROM & Reserved, zero \\
   \hex{4000} & \hex{7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
-  \hex{8000} & \hex{FFFF} & ROM & Read-only (operating system) \\
+  \hex{8000} & \hex{FFFF} & ROM & Operating system \\
   \hline

### 2018-07-03 paper.tex
SWAP, RULE, and RAND

@@
 % Memory address
+\newcommand\code[1]{{\tt #1}}
 \newcommand\hex[1]{{\tt #1}}
@@
 
-Each cell has 256 bytes (one page) of state.
+Each cell has 256 bytes (one page) of volatile state, and a 32K read-only operating system.
 All cells' bytes are initially zero, unless otherwise specified
@@
 mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
-Thus, in a memory-mapped 16-bit address, the bits have the meanings shown in Table~\ref{tab:AddressBits}.
+Thus, in a memory-mapped 16-bit address, the bits have the interpretation shown in Table~\ref{tab:AddressBits}.
 Note the exception: cell $(x,y,z)$ is not mapped to page 85 (\hexrange{5500}{55FF})
@@
 is copied back to the cellular automata storage.
+
+\subsection{Special addresses}
+
+There are several emulator-mapped memory addresses that perform special functions:
+the write-only addresses \code{SWAP} (\hex{FFF7}) and \code{RULE} (\hex{FFF8}),
+and the read-only address \code{RAND} (\hex{FFF9}).
+
+Any value written to \code{SWAP} or \code{RULE} is interpreted as the index of a neighboring cell.
+Only the lower 6 bits of the written byte are relevant, and specify a cell in the $3 \times 3 \times 3$ Moore neighborhood
+in the same way as bits 8-13 of a memory-mapped storage address (c.f. Table~\ref{tab:AddressBits}).
+
+Writing a byte to \code{SWAP} has the effect of swapping the storage and operating system pages associated with the current cell and the specified neighboring cell,
+and then terminating execution.
+If the neighboring cell is outside the CA (see Section~\ref{sec:AdjacentCells}),
+this updates the current cell (and its operating system) but has no effect on the neighboring cell.
+
+Writing a byte to \code{RULE} copies the current cell's operating system to the specified neighboring cell
+(unless the neighboring cell is outside the CA, in which case it has no effect).
+
+Reading a byte from \code{RAND} yields 8 random bits, newly sampled with each read (i.e. multiple reads yield different results).
 
@@
 to discourage frivolous edits.
-(Suggest $W=1$ to keep this simple.
-Minting a coin lets you write a cell.
-Seems fair.)
+If (for example) $W=1$, then minting a coin lets you write a cell.
+
+As well as writing the (volatile) state of the cell, you can write its 32K operating system
+(mapped to address range \hexrange{8000}{FFEF} in the virtual machine; the last 16 bytes, corresponding to \hexrange{FFF0}{FFFF}, are ignored).
+After the first time an operating system is written, it can be referred to by SHA-256 hash, to save space.
 
@@
   \hex{0000} & \hex{00FF} & RAM (storage-mapped) & State of cell $(x,y,z)$ \\
-  \hex{0100} & \hex{01FF} & RAM (transient) & Stack, initialized to zero \\
-  \hex{0200} & \hex{03FF} & RAM (transient) & Scratch memory, initialized to zero \\
-  \hex{0400} & \hex{3FFF} & ROM & Reserved, zero \\
+  \hex{0100} & \hex{01FF} & RAM (transient) & Stack: initialized to zero \\
+  \hex{0200} & \hex{03FF} & RAM (transient) & Scratch memory: initialized to zero \\
+  \hex{0400} & \hex{3FFF} & ROM & Reserved: zero \\
   \hex{4000} & \hex{7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
-  \hex{8000} & \hex{FFFF} & ROM & Operating system \\
+  \hex{8000} & \hex{FFEF} & ROM & Operating system \\
+  \hex{FFF0} & \hex{FFF6} & ROM & Reserved: zero \\
+  \hex{FFF7} & \hex{FFF7} & Write-only & \code{SWAP}: swap storage and O/S pages \\
+  \hex{FFF8} & \hex{FFF8} & Write-only & \code{RULE}: copy O/S pages to neighbor \\
+  \hex{FFF9} & \hex{FFF9} & Read-only & \code{RAND}: random byte generator \\
+  \hex{FFFA} & \hex{FFFF} & ROM & Reset and interrupt vectors: zero \\
   \hline
@@
 \begin{table}
-\begin{tabular}{ll}
+\begin{tabular}{lll}
   \hline
-  Bit(s) & Meaning \\
+  Byte & Bit(s) & Meaning \\
   \hline
-  14,15 & Must be \hex{01} \\
-  12,13 & $i$ \\
-  10,11 & $j$ \\
-  8,9 & $k$ \\
-  0-7 & Address within page \\
+  MSB & 4,5 & $i$ \\
+  MSB & 2,3 & $j$ \\
+  MSB & 0,1 & $k$ \\
+  LSB & 0-7 & Address within page \\
   \hline
@@
 \caption{
-  Interpretation of an address in the memory-mapped range \hexrange{4000}{7FFF}
-  as pointing to a cell $(x+i-1,y+j-1,z+k-1)$.
+  Interpretation of a 16-bit address in the memory-mapped range \hexrange{4000}{7FFF}
+  as pointing to the cell $(x+i-1,y+j-1,z+k-1)$.
   See Table~\ref{tab:MooreNeighborhood} and Table~\ref{tab:VonNeumannNeighborhood}.

### 2018-07-03 paper.tex
O/S

@@
 
-Each cell has 256 bytes (one page) of volatile state, and a 32K read-only operating system.
+Each cell has 256 bytes (one page) of volatile state.
 All cells' bytes are initially zero, unless otherwise specified
@@
 These bytes are the storage of the virtual machine.
+
+Each cell also has a 32-kilobyte read-only operating system.
+Most cells share the same operating system, and operating system changes should be rare,
+so the O/S for a cell can be represented using a compact hash key such as SHA-256.
 
@@
 There are several emulator-mapped memory addresses that perform special functions:
-the write-only addresses \code{SWAP} (\hex{FFF7}) and \code{RULE} (\hex{FFF8}),
-and the read-only address \code{RAND} (\hex{FFF9}).
+the write-only address \code{SWAP} (\hex{FFF8}), which facilitates diffusion models,
+and the read-only address \code{RAND} (\hex{FFF9}), which provides a random number generator.
 
-Any value written to \code{SWAP} or \code{RULE} is interpreted as the index of a neighboring cell.
+Any value written to \code{SWAP} is interpreted as the index of a neighboring cell.
 Only the lower 6 bits of the written byte are relevant, and specify a cell in the $3 \times 3 \times 3$ Moore neighborhood
@@
 
-Writing a byte to \code{RULE} copies the current cell's operating system to the specified neighboring cell
-(unless the neighboring cell is outside the CA, in which case it has no effect).
+Writes to \code{SWAP} are not the only means by which operating system code is propagated across the CA.
+During an update, writing any byte to any neighboring cell, i.e. any cell in the memory-mapped storage (\hexrange{4000}{7FFF}),
+has the implicit effect of copying the operating system pages from the current cell to the neighbor.
+This is handled transparently by the emulator without consuming any additional cycles.
 
@@
 As well as writing the (volatile) state of the cell, you can write its 32K operating system
-(mapped to address range \hexrange{8000}{FFEF} in the virtual machine; the last 16 bytes, corresponding to \hexrange{FFF0}{FFFF}, are ignored).
-After the first time an operating system is written, it can be referred to by SHA-256 hash, to save space.
+(mapped to address range \hexrange{8000}{FFF7} in the virtual machine; the last 16 bytes, corresponding to \hexrange{FFF8}{FFFF}, are ignored).
+After the first time an operating system is used in the blockchain, it can be referred to by SHA-256 hash, to save space.
 
@@
   \hex{4000} & \hex{7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
-  \hex{8000} & \hex{FFEF} & ROM & Operating system \\
-  \hex{FFF0} & \hex{FFF6} & ROM & Reserved: zero \\
-  \hex{FFF7} & \hex{FFF7} & Write-only & \code{SWAP}: swap storage and O/S pages \\
-  \hex{FFF8} & \hex{FFF8} & Write-only & \code{RULE}: copy O/S pages to neighbor \\
+  \hex{8000} & \hex{FFF7} & ROM & Operating system \\
+  \hex{FFF8} & \hex{FFF8} & Write-only & \code{SWAP}: swap storage page with neighbor \\
   \hex{FFF9} & \hex{FFF9} & Read-only & \code{RAND}: random byte generator \\

### 2018-07-03 paper.tex
O/S vectors

@@
 As well as writing the (volatile) state of the cell, you can write its 32K operating system
-(mapped to address range \hexrange{8000}{FFF7} in the virtual machine; the last 16 bytes, corresponding to \hexrange{FFF8}{FFFF}, are ignored).
+(mapped to address range \hexrange{8000}{FFF7} in the virtual machine; the last 8 bytes, corresponding to \hexrange{FFF8}{FFFF}, are ignored).
 After the first time an operating system is used in the blockchain, it can be referred to by SHA-256 hash, to save space.

### 2018-07-03 paper.tex
ok

@@
 \begin{abstract}
-  As a toy illustration,
-  this paper describes a blockchain for cellular automata
-  comprising a 3D array of virtual machines.
-  Each machine emulates a 6502 processor with access to one page (256 bytes) of persistent storage,
+  As a toy example of incentivized, distributed spatial computation,
+  this paper develops a blockchain for cellular automata
+  comprising a 3D array of virtual 6502 processors, each with access to one page (256 bytes) of persistent storage,
   as well as the pages of its neighbors (and a read-only operating system).
@@
 
-\section{Cellular automata}
+\section{Introduction}
 
-\subsection{Storage}
-\label{sec:Storage}
+Blockchains are useful for cryptocurrencies.
+They can also be used to drive computation on a distributed virtual machine.
+In the Ethereum network, this is used to implement smart contracts.
+Here we describe a different use: the simulation of a virtual world of general cellular automata.
 
-The cellular automaton (CA) is a $L \times L \times L$ array of cells,
+The cellular automata we develop are both universal and familiar, consisting of a three-dimensional grid of 6502 microprocessors broken into blocks.
+The blockchain framework incentivizes simulation of individual cell blocks, as well as data-sharing between blocks.
+
+This framework is intended for recreational purposes: the coins mined by the system
+have no ostensible purpose other than regulating the rate at which participants can write to the simulated world.
+
+\section{Methods}
+
+\subsection{Cell blocks}
+
+Every cell of the cellular automata has a unique integer 3-space co-ordinate $(x,y,z)$.
+
+A single {\bf cell block} of the cellular automata is a $L \times L \times L$ array of cells,
 located at an offset $(\alpha_x,\alpha_y,\alpha_z)$ in integer 3-space.
-Cells in the CA are indexed $(x,y,z$) where
+Cells in the cell block are indexed $(x,y,z$) where
 $\alpha_x \leq x < \alpha_x+L$,
@@
 
-Each cell has 256 bytes (one page) of volatile state.
+\subsection{Read-only and random-access storage}
+\label{sec:Storage}
+
+In designing general cellular automata running virtual machines,
+it is useful to distinguish between a cell's read-only memory (ROM)
+and its random-access memory (RAM).
+The former is useful for storing the cell's operating program (or immutable parts of it); the latter, for tracking its internal state.
+As an analogy, programs that don't use the operating system are, by necessity, small and simple (c.f. cells without ROM);
+programs that don't use variables may be inflexible (c.f. cells without RAM).
+
+Each cell has 256 bytes (one page) of updatable state (RAM).
 All cells' bytes are initially zero, unless otherwise specified
-(including, implicitly, cells whose coordinates are outside the CA; see also Section~\ref{sec:AdjacentCells}).
+(including, implicitly, cells whose coordinates are outside the cell block; see also Section~\ref{sec:AdjacentCells}).
 These bytes are the storage of the virtual machine.
 
-Each cell also has a 32-kilobyte read-only operating system.
+Each cell also has a 32-kilobyte read-only operating system (ROM).
 Most cells share the same operating system, and operating system changes should be rare,
@@
 
+\subsection{Updating the cell block}
+\label{sec:UpdatingCellBlock}
+
+An {\bf update} of the cell block consists of updating each cell in a pseudorandom order.
+This can be any order that visits $L^3$ uniformly-random selected cells in a well-defined pseudorandom order;
+a Mersenne Twister (MT19937),
+seeded on the tuple $(\alpha_x,\alpha_y,\alpha_z,L,U)$ where $U$ is the number of updates,
+is a suitable random number generator (RNG) for this and other random numbers required during the update.
+
 \subsection{Virtual machine}
 
-An {\bf update} of the CA consists of updating each cell in the CA in a pseudorandom order.
-This can be any order that visits $L^3$ uniformly-random selected cells;
-a Mersenne Twister is suitable for this.
+A cell $(x,y,z)$ is updated by emulating a virtual machine.
+In designing this virtual machine we want to use a well-known, simple architecture with small programs.
+The interface between the virtual machine and the cell block
+needs to be such that the cell's ROM and the RAM, as well as the RAM of its neighbors,
+are accessible to the code running in the virtual machine.
 
-A cell $(x,y,z)$ is updated by emulating an 8-bit virtual machine.
-The CPU is the MOS Technology 6502 processor.
-The memory map is shown in Table~\ref{tab:MemoryMap}.
+After some experiments with 32-bit ARM RISC architectures,
+I settled on the 8-bit MOS Technology 6502 processor for a CPU.
+The corresponding 16-bit memory map is shown in Table~\ref{tab:MemoryMap}.
+
 In this memory map, the 27 pages of storage
@@
 in the $3 \times 3 \times 3$ unit cube centered on $(x,y,z)$
+(in cellular automata terminology, the {\em three-dimensional Moore neighborhood})
 are mapped directly to bytes in the address range \hexrange{4000}{7FFF} of the virtual machine.
-
-The detailed description of the memory map for this $3 \times 3 \times 3$ neighborhood,
-which in cellular automata terminology is
-the {\em three-dimensional Moore neighborhood},
+The detailed description of this part of the memory map
 is given in Table~\ref{tab:MooreNeighborhood}.
 
-The neighborhood memory map can be understood as the cell $(x+i-1,y+j-1,z+k-1)$
+The Moore neighborhood memory map can be understood as the cell $(x+i-1,y+j-1,z+k-1)$
 mapping to page number (64+i*16+j*4+k) with $i,j,k \in \{0,1,2\}$.
@@
 because it is mapped to zero page (\hexrange{0000}{00FF}) instead.
+
+\subsection{Timeline of a cell update}
+\label{sec:Timeline}
 
@@
 or by non-maskable interrupt after
-completion of the first instruction that takes the cycle count past 65,536 cycles.
+completion of the first instruction that takes the cycle count past {\tt STOPTIME} cycles,
+where {\tt STOPTIME} is a 16-bit integer.
+For notes on how {\tt STOPTIME} is chosen, see Section~\ref{sec:Timeouts}.
 
@@
 
+\subsection{Timeouts}
+\label{sec:Timeouts}
+
+To encourage quicker programs,
+we make the number of cycles available to each update a random variable, instead of a fixed limit.
+
+The random variable {\tt STOPTIME} is generated using the cell block RNG (Section~\ref{sec:UpdatingCellBlock})
+in the following way:
+\begin{itemize}
+\item Sample $m$ from the following (bounded geometric) distribution
+  \[
+  P(m=k) = \left\{ \begin{array}{ll} 2^{1-k} & \mbox{if $2 \leq k < 15$} \\ 2^{-13} & \mbox{if $k = 15$} \end{array} \right.
+    \]
+  \item Let $r$ denote a random 16-bit integer
+  \item Set {\tt STOPTIME} to $(2^m \vee (r \wedge (2^m - 1))$
+\end{itemize}
+
+The mean value of {\tt STOPTIME} is 44.5 cycles,
+but there is a small ($\sim$0.5\%) chance that it's over 1,000 cycles,
+and an even smaller ($\sim$0.05\%) chance that it's over 10,000 cycles.
+
 \subsection{Special addresses}
@@
 There are several emulator-mapped memory addresses that perform special functions:
-the write-only address \code{SWAP} (\hex{FFF8}), which facilitates diffusion models,
-and the read-only address \code{RAND} (\hex{FFF9}), which provides a random number generator.
+the read-only \code{TIME} (\hexrange{FFF6}{FFF7}), which gives the time until termination;
+the write-only \code{SWAP} (\hex{FFF8}), which facilitates diffusion models;
+and the read-only \code{RAND} (\hex{FFF9}), which provides a random number generator.
 
+These are described in the sections below.
+
+\subsection{Operating system updates}
+
+The machine architecture described so far is adequate if the operating system ROM is shared by every cell on the block.
+To allow O/S updates to be propagated across the block, we introduce two specific mechanisms.
+Guiding the design of this mechanism are the following two principles:
+\begin{itemize}
+\item If a cell writes another cell's RAM, then it should expect that cell's ROM to be the same as its own.
+\item Cells should, nevertheless, be allowed to diffuse past each other without necessarily overwriting each others' ROM.
+\end{itemize}
+
+The first mechanism for O/S propagation is the write-only \code{SWAP} address, \hex{FFF8}.
 Any value written to \code{SWAP} is interpreted as the index of a neighboring cell.
@@
 and then terminating execution.
-If the neighboring cell is outside the CA (see Section~\ref{sec:AdjacentCells}),
+If the neighboring cell is outside the cell block (see Section~\ref{sec:AdjacentCells}),
 this updates the current cell (and its operating system) but has no effect on the neighboring cell.
 
-Writes to \code{SWAP} are not the only means by which operating system code is propagated across the CA.
+Reading a byte from address \code{SWAP} always yields zero, and has no other effect.
+
+As mentioned above, the \code{SWAP} interface is one of two mechanisms by which operating system code is propagated across the cell block.
+The second mechanism is built into the memory-mapped storage interface itself.
 During an update, writing any byte to any neighboring cell, i.e. any cell in the memory-mapped storage (\hexrange{4000}{7FFF}),
@@
 
+\subsection{Random numbers}
+
+The read-only \code{RAND} address, \hex{FFF9}, is a random number generator.
 Reading a byte from \code{RAND} yields 8 random bits, newly sampled with each read (i.e. multiple reads yield different results).
+The random numbers are provided by the cell block's RNG (Section~\ref{sec:UpdatingCellBlock}).
+
+\subsection{Timer}
+
+The read-only 16-bit \code{TIME} address, \hexrange{FFF6}{FFF7}, gives the number of cycles before the program is terminated.
+(More precisely, this address is initialized to \code{STOPTIME} before the first instruction is executed,
+and decremented after completion of each instruction.)
 
@@
 
-Consider those cell coordinates $(x,y,z)$ that are outside the CA,
+Consider those cell coordinates $(x,y,z)$ that are outside the cell block,
 e.g. $(\alpha_x-1,\alpha_y,\alpha_z)$ or  $(\alpha_x,\alpha_y,\alpha_z + L)$.
-Some of these, specifically those in the Moore neighborhood of cells that are in the CA,
+Some of these, specifically those in the Moore neighborhood of cells that are in the cell block,
 are a relevant initial condition for the machine,
@@
 
-We refer to these cells as the CA's {\bf adjacent cells}.
+We refer to these cells as the cell block's {\bf adjacent cells}.
 Earlier, in Section~\ref{sec:Storage}, we specified that the state of these cells was implicitly initialized to zero.
 More generally, we may refer to the initial (and perpetual, throughout the execution of an update)
-state of these cells as the CA's {\bf adjacent state}.
+state of these cells as the cell block's {\bf adjacent state}.
 This adjacent state is zero until otherwise specified, but may be affected
-by interactions with neighboring CAs in the infinite vector space,
+by interactions with neighboring cell blocks in the infinite vector space,
 as for example by the blockchain of Section~\ref{sec:Blockchain}.
 
-\section{Blockchain}
+\subsection{Blockchain}
 \label{sec:Blockchain}
 
-The CA has an associated blockchain, defined as in \cite{Nakamoto2008}.
+The cell block has an associated blockchain, defined as in \cite{Nakamoto2008}.
 Updates are incentivized by mining,
-and the reward for mining is adjusted to include redemption of coins mined by adjacent CA blockchains
-(this also updates the adjacent cell states accounted for by those CAs).
+and the reward for mining is adjusted to include redemption of coins mined by adjacent cell-block blockchains
+(this also updates the adjacent cell states accounted for by those cell blocks).
 
-\subsection{Updates}
+\subsection{Driving updates with the blockchain}
 
-A valid block must include the $256 \times L^3$ bytes describing the state of the CA following $C$ updates
+A valid block must include the $256 \times L^3$ bytes describing the state of the cell block following $C$ updates
 from the state at the end of the previous block.
@@
 
-Let the definition of a block include the coordinates $(\alpha_x,\alpha_y,\alpha_z,L)$;
-these coordinates, together with the number of updates,
-must be used to seed the random number generator for the $Z$ updates.
+Let the definition of a block include the coordinates $(\alpha_x,\alpha_y,\alpha_z,L,U)$
+used to seed the cell block's random number generator and identify the cell block in space and time.
 
@@
 Any transaction in the blockchain may include, as additional data,
-a {\bf write operation} which modifies the state of one or more cells of the CA.
+a {\bf write operation} which modifies the state of one or more cells of the cell block.
 This modification is considered to take place after the emulation of $Z$ update cycles
@@
 As well as writing the (volatile) state of the cell, you can write its 32K operating system
-(mapped to address range \hexrange{8000}{FFF7} in the virtual machine; the last 8 bytes, corresponding to \hexrange{FFF8}{FFFF}, are ignored).
+(mapped to address range \hexrange{8000}{FFEF} in the virtual machine; the last 16 bytes, corresponding to \hexrange{FFF0}{FFFF}, are ignored).
 After the first time an operating system is used in the blockchain, it can be referred to by SHA-256 hash, to save space.
@@
 The block may include sections of blockchain from {\bf adjacent cellular automata},
-defined to be CAs with co-ordinate ranges that
-overlap with the current CA's adjacent cells,
+defined to be cell blocks with co-ordinate ranges that
+overlap with the current cell block's adjacent cells,
 but no cells that overlap with
-cells in the co-ordinate range of this CA (or any other CA previously encountered in the blockchain,
-except the adjacent CA itself).
-These blockchain segments from adjacent CAs are called {\bf adjacent segments}.
+cells in the co-ordinate range of this cell block (or any other cell block previously encountered in the blockchain,
+except the adjacent cell block itself).
+These blockchain segments from adjacent cell blocks are called {\bf adjacent segments}.
 
 Adjacent segments must be consistent with any previously-observed adjacent segments
-earlier on the current blockchain (that is, they must not involve conflicting CA,
-and they must otherwise be consistent continuations of the adjacent CA's blockchain).
+earlier on the current blockchain (that is, they must not involve conflicting cell block,
+and they must otherwise be consistent continuations of the adjacent cell block's blockchain).
 They may, themselves, recursively refer to parts of the current blockchain as their own adjacent segments,
@@
 
-When a block includes an adjacent segment, the CA's adjacent cell state is updated to match the adjacent segment.
-Regular mutual inclusion of adjacent segments from adjacent CAs' blockchains
-corresponds to keeping a bidirectional read-only interface open between those CAs.
+When a block includes an adjacent segment, the cell block's adjacent cell state is updated to match the adjacent segment.
+Regular mutual inclusion of adjacent segments from adjacent cell blocks' blockchains
+corresponds to keeping a bidirectional read-only interface open between those cell blocks.
 
@@
 where $C$ is the number of {\bf redeemable coins} mined or accrued (via more distant, proximal adjacencies)
-by cells in the adjacent CA,
-and $R$ is the {\bf redemption rate}, the number of redeemable coins in adjacent CAs
-required to accrue a coin in the current CA.
+by cells in the adjacent cell block,
+and $R$ is the {\bf redemption rate}, the number of redeemable coins in adjacent cell blocks
+required to accrue a coin in the current cell block.
 It is of course necessary to have $R>2$ to prevent an explosion of redeemable coins in 3D space;
@@
 
-Setting the parameters $L$ (CA size) and $C$ (updates per block)
+\subsection{Parameters}
+
+Setting the parameters $L$ (cell block size) and $C$ (updates per block)
 is a little arbitrary, but can be guided by target memory footprint and clock rate.
-For memory, the CA state requires $256 \times L^3$ bytes.
-For clock rate, if we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of $2^{15}$ cycles per cellular update,
-we would update the whole board $10^6 \times T / (2^{15} \times L^3)$ times
+For memory, the cell block state requires $256 \times L^3$ bytes.
+For clock rate, if we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of $44.5$ cycles per cellular update,
+we would update the whole cell block $10^6 \times T / (44.5 \times L^3)$ times
 where $T$ is the time in seconds between blocks ($T=36,000$ for bitcoin).
@@
 each cell is updated every few minutes;
-each block takes at most 64Mb to describe the CA state;
+each block takes at most 64Mb to describe the cell block state;
 and the computational burden is roughly comparable to a 6502 CPU.
@@
   \hex{4000} & \hex{7FFF} & RAM (storage-mapped) & Memory-mapped neighborhood (Table~\ref{tab:MooreNeighborhood}) \\
-  \hex{8000} & \hex{FFF7} & ROM & Operating system \\
+  \hex{8000} & \hex{FFF0} & ROM & Operating system \\
+  \hex{FFF0} & \hex{FFF5} & ROM & Reserved; zero \\
+  \hex{FFF6} & \hex{FFF7} & Read-only & \code{TIME}: number of cycles left until termination \\
   \hex{FFF8} & \hex{FFF8} & Write-only & \code{SWAP}: swap storage page with neighbor \\
@@
   writing to these addresses has no effect.
-  Pages mapping to cells outside the CA (e.g. $x<\alpha_x$ or $x \geq \alpha_x+L$, or similarly for $y,z$)
+  Pages mapping to cells outside the cell block (e.g. $x<\alpha_x$ or $x \geq \alpha_x+L$, or similarly for $y,z$)
   are read-only and zero (unless otherwise specified, as in Section~\ref{sec:AdjacentCells}).

### 2018-07-03 paper.tex
ok

@@
 
-Blockchains are useful for cryptocurrencies.
-They can also be used to drive computation on a distributed virtual machine.
+Cellular Automata (CA) are a common model for universal computation.
+The idea that every pixel (or voxel) of the world computes and talks to its neighbors
+is an appealing synthesis of concepts from natural science such as time- and space-invariance, and wave propagation.
+Cellular automata (along with agent models, lattice gases, and other discretized frameworks) have been useful models in physics and biology,
+and have also been used in computer games and artificial life experiments (A-Life).
+
+Although straightforward to implement, cellular automata experiments are limited by the resources available to an individual;
+furthermore, there are no well-established ways to bridge different experiments (e.g. to make a link between separate CA worlds).
+Here, we describe a way to do this using the distributed timestamp service provided by a proof-of-work blockchain.
+
+The blockchain was introduced for cryptocurrencies \cite{Nakamoto2008}.
+It can also be used to drive computation on a distributed virtual machine.
 In the Ethereum network, this is used to implement smart contracts.
-Here we describe a different use: the simulation of a virtual world of general cellular automata.
+Here we describe a different (simpler) use: the simulation of a virtual world of generic cellular automata.
 
-The cellular automata we develop are both universal and familiar, consisting of a three-dimensional grid of 6502 microprocessors broken into blocks.
+The cellular automata we develop as a motivating example are both universal and familiar, consisting of a three-dimensional grid of 6502 microprocessors broken into blocks.
 The blockchain framework incentivizes simulation of individual cell blocks, as well as data-sharing between blocks.
 
-This framework is intended for recreational purposes: the coins mined by the system
+I chose the 6502 processor as the virtual machine for several reasons.
+First, I like it, and so do many others, which is probably why it still gets attention 40 years after it appeared \cite{JonasKording2017}.
+Second, it's simple, and compact (a slim memory footprint per cell is important for a CA; for example, 32 bits per instruction is too much).
+Why not use RedCode, the programming language of Core War \cite{Dewdney1984}?
+Core War is well-established among afficionados, and is studied by A-Life researchers \cite{VowkWaitSchmidt2004},
+but it's poorly suited to spatially mapped simulations, since Core War programs run in a linear memory space.
+Other prior A-Life frameworks have used virtual machines derived from 80x86 processors, such as Avida \cite{AdamiBrown1994}.
+To define a system with enough precision that it can be implemented on a blockchain, it's helpful to stay close to a machine that actually exists,
+so we use 6502 with only a few minimal extra functions that are all memory-mapped, rather like hardware might be on a real 6502 system.
+
+The blockchain framework developed here is intended for recreational purposes: the coins mined by the system
 have no ostensible purpose other than regulating the rate at which participants can write to the simulated world.
+They might play the role of a score in a game world, measuring the contribution of the owner to keeping the world running.
+However, alternate scores (metrics) of success are possible in the world we describe, such as the popularity or success of programs that one has introduced into the world.
 
@@
 The former is useful for storing the cell's operating program (or immutable parts of it); the latter, for tracking its internal state.
-As an analogy, programs that don't use the operating system are, by necessity, small and simple (c.f. cells without ROM);
-programs that don't use variables may be inflexible (c.f. cells without RAM).
+
+We allow cells to have both types of memory, RAM (for mutable state) and ROM (for immutable behavior).
+It is perfectly possible to write cells with only one type of memory.
+As an analogy, automata without ROM are like programs that run in user space without calling the operating system; they are by necessity small and simple.
+Automata without RAM are like stateless protocols: they scale well, but they may be inflexible for some purposes.
 
@@
 A cell $(x,y,z)$ is updated by emulating a virtual machine.
-In designing this virtual machine we want to use a well-known, simple architecture with small programs.
+In designing this virtual machine we want to use a well-known, simple architecture with small programs,
+that is transparently universal.
 The interface between the virtual machine and the cell block
@@
 At the beginning of an update,
-the registers {\tt A,X,Y} are initialized to zero,
-all status flags (register {\tt P}) are set to zero,
-the stack pointer (register {\tt SP}) is set to \hex{FF},
-the program counter {\tt PC} is set to \hex{0000},
+the registers \code{A,X,Y} are initialized to zero,
+all status flags (register \code{P}) are set to zero,
+the stack pointer (register \code{SP}) is set to \hex{FF},
+the program counter \code{PC} is set to \hex{0000},
+the cellular automata states are copied into memory (to zero page and other pages shown in Table~\ref{tab:MooreNeighborhood}),
 and execution begins at the first byte of zero page.
@@
 when a software interrupt or undefined instruction is encountered,
-or by non-maskable interrupt after
-completion of the first instruction that takes the cycle count past {\tt STOPTIME} cycles,
-where {\tt STOPTIME} is a 16-bit integer.
-For notes on how {\tt STOPTIME} is chosen, see Section~\ref{sec:Timeouts}.
+or by non-maskable interrupt after completion of the first instruction that takes the cycle count past \code{STOPTIME} cycles,
+where \code{STOPTIME} is a 16-bit integer.
+For notes on how \code{STOPTIME} is chosen, see Section~\ref{sec:Timeouts}.
 
 On termination, the transient address ranges in the memory map
-(including the stack)
-are discarded.
-The neighborhood memory map (\hexrange{4000}{7FFF}, Table~\ref{tab:MooreNeighborhood})
-is copied back to the cellular automata storage.
+(including the stack) are discarded.
+If execution was terminated by timeout, then the rest of memory is also discarded,
+and the state of the CA is restored
+Otherwise, if the program exited (by software interrupt or undefined opcode) before timing out,
+the neighborhood memory-mapped pages (zero page and other pages in the range \hexrange{4000}{7FFF}, see Table~\ref{tab:MooreNeighborhood})
+are copied back to the cellular automata storage.
 
@@
 
+We need to set a limit on how many cycles a program is allowed to run for (i.e. \code{STOPTIME}).
+The problem is that, if we set a hard limit, randomly emergent programs will tend to hit that limit
+(e.g. by entering infinite loops)
+and human programmers will be tempted to surf close to it.
+
 To encourage quicker programs,
-we make the number of cycles available to each update a random variable, instead of a fixed limit.
+we make \code{STOPTIME} a random variable, instead of a fixed limit.
+We also require that programs exit gracefully (returning control to the emulator via a software interrupt)
+if their changes are to be committed to storage.
+As well as discouraging slow programs, this makes it easier to write programs that are atomic (all-or-nothing) in their effects.
+Programs can check the \code{TIME} timer at \hexrange{FFF6}{FFF7} to see how many cycles are available.
 
-The random variable {\tt STOPTIME} is generated using the cell block RNG (Section~\ref{sec:UpdatingCellBlock})
-in the following way:
+The random variable \code{STOPTIME} is generated using the cell block RNG (Section~\ref{sec:UpdatingCellBlock})
+and the following algorithm:
 \begin{itemize}
-\item Sample $m$ from the following (bounded geometric) distribution
+\item Let $r$ and $s$ denote random 16-bit integers
+\item Let $z$ be the number of leading zero bits of $s$, so $0 \leq z \leq 16$
+\item Let $m = \mbox{max}(z,13) + 2$. The distribution of $m$ is now given by
   \[
@@
   \]
-  \item Let $r$ denote a random 16-bit integer
-  \item Set {\tt STOPTIME} to $(2^m \vee (r \wedge (2^m - 1))$
+\item Set \code{STOPTIME} to $(2^m \vee (r \wedge (2^m - 1))$
 \end{itemize}
 
-The mean value of {\tt STOPTIME} is 44.5 cycles,
+% This Perl one-liner calculates the mean value of STOPTIME and some related probabilities
+% perl -e '$e=0;for($m=2;$m<=15;++$m){$p=($m<15?(2**(1-$m)):(2**(-13)));$b=2**$m;$s=($b+($b-1)/2);print"$p $t $s\n";$e+=$p*$s;$t+=$p}print"$t $e\n"'
+
+The mean value of \code{STOPTIME} is 44.5 cycles,
 but there is a small ($\sim$0.5\%) chance that it's over 1,000 cycles,
@@
 There are several emulator-mapped memory addresses that perform special functions:
-the read-only \code{TIME} (\hexrange{FFF6}{FFF7}), which gives the time until termination;
+the read-only \code{TIME} (\hexrange{FFF6}{FFF7}), which is the number of cycles until forced termination;
 the write-only \code{SWAP} (\hex{FFF8}), which facilitates diffusion models;
@@
 Writing a byte to \code{SWAP} has the effect of swapping the storage and operating system pages associated with the current cell and the specified neighboring cell,
-and then terminating execution.
+and then triggering a software interrupt, terminating execution.
 If the neighboring cell is outside the cell block (see Section~\ref{sec:AdjacentCells}),
 this updates the current cell (and its operating system) but has no effect on the neighboring cell.
+If the written byte is not a valid index byte, then there is no effect.
+A software interrupt is triggered, and program execution terminated, regardless of whether the written byte is a valid index pointing to a cell within the cell block.
 
@@
 \newcommand\memrow[4]{
-    {\tt {#1}00} & {\tt {#1}FF} & State of cell $(x#2,y#3,z#4)$ \\
+    \hex{{#1}00} & \hex{{#1}FF} & State of cell $(x#2,y#3,z#4)$ \\
 }

### 2018-07-03 paper.tex
cube

@@
 \documentclass{article}
+
+\usepackage{graphicx}
 
@@
 
-The blockchain was introduced for cryptocurrencies \cite{Nakamoto2008}.
-It can also be used to drive computation on a distributed virtual machine.
-In the Ethereum network, this is used to implement smart contracts.
+The blockchain was introduced as a distributed ledger for cryptocurrencies \cite{Nakamoto2008}.
+More generally it can be used to drive computation on a distributed virtual machine.
+In the Ethereum network, the virtual machine is used to implement smart contracts.
 Here we describe a different (simpler) use: the simulation of a virtual world of generic cellular automata.
@@
 I settled on the 8-bit MOS Technology 6502 processor for a CPU.
-The corresponding 16-bit memory map is shown in Table~\ref{tab:MemoryMap}.
+The corresponding 16-bit memory map is shown in Table~\ref{tab:MemoryMap} and Figure~\ref{fig:UnitCube}.
 
@@
 in the $3 \times 3 \times 3$ unit cube centered on $(x,y,z)$
-(in cellular automata terminology, the {\em three-dimensional Moore neighborhood})
+(in cellular automata terminology, the {\em three-dimensional Moore neighborhood}, shown in Figure~\ref{fig:UnitCube})
 are mapped directly to bytes in the address range \hexrange{4000}{7FFF} of the virtual machine.
@@
   as pointing to the cell $(x+i-1,y+j-1,z+k-1)$.
-  See Table~\ref{tab:MooreNeighborhood} and Table~\ref{tab:VonNeumannNeighborhood}.
+  See Table~\ref{tab:MooreNeighborhood} and Table~\ref{tab:VonNeumannNeighborhood}, and Figure~\ref{fig:UnitCube}.
   Note that $i,j,k, \in \{ 0,1,2 \}$.
@@
 \end{table}
+
+\begin{figure}
+  \includegraphics[width=\textwidth]{unitcube.pdf}
+\caption{
+  The unit cube of the Moore neighborhood, and its relationship to the memory map of Table~\ref{tab:MooreNeighborhood}.
+  The central cell is at $(1,1,1)$, and is mapped to zero page.
+  All other cells are mapped to pages in the range \hexrange{4000}{7FFF},
+  the first six bits of the address MSB corresponding to the $x$, $y$ and $z$ co-ordinates within this cell (see Table~\ref{tab:AddressBits}).
+  \label{fig:UnitCube}
+}
+\end{figure}
 

### 2018-07-03 paper.tex
update

@@
 is an appealing synthesis of concepts from natural science such as time- and space-invariance, and wave propagation.
-Cellular automata (along with agent models, lattice gases, and other discretized frameworks) have been useful models in physics and biology,
+Cellular automata, along with agent models, lattice gases, and other discretized frameworks, have been useful models in physics and biology,
 and have also been used in computer games and artificial life experiments (A-Life).
@@
 Although straightforward to implement, cellular automata experiments are limited by the resources available to an individual;
-furthermore, there are no well-established ways to bridge different experiments (e.g. to make a link between separate CA worlds).
+furthermore, there are no well-established ways to bridge different experiments, for example to make a link between separate CA worlds.
 Here, we describe a way to do this using the distributed timestamp service provided by a proof-of-work blockchain.
-
-The blockchain was introduced as a distributed ledger for cryptocurrencies \cite{Nakamoto2008}.
-More generally it can be used to drive computation on a distributed virtual machine.
-In the Ethereum network, the virtual machine is used to implement smart contracts.
-Here we describe a different (simpler) use: the simulation of a virtual world of generic cellular automata.
 
@@
 The blockchain framework incentivizes simulation of individual cell blocks, as well as data-sharing between blocks.
+
+\subsection{Why 6502s?}
 
@@
 but it's poorly suited to spatially mapped simulations, since Core War programs run in a linear memory space.
-Other prior A-Life frameworks have used virtual machines derived from 80x86 processors, such as Avida \cite{AdamiBrown1994}.
-To define a system with enough precision that it can be implemented on a blockchain, it's helpful to stay close to a machine that actually exists,
-so we use 6502 with only a few minimal extra functions that are all memory-mapped, rather like hardware might be on a real 6502 system.
+Other prior A-Life frameworks have used virtual machines derived from 80x86 processors, as in Avida \cite{AdamiBrown1994}.
 
-The blockchain framework developed here is intended for recreational purposes: the coins mined by the system
+To define a new virtual machine with sufficient precision, it's helpful to stay close to a machine that physically exists.
+We use 6502s with a very small set of machine extensions that are all memory-mapped, rather like hardware might be on a real 6502 system.
+
+\subsection{Why blockchains?}
+
+The blockchain was introduced as a distributed ledger for cryptocurrencies \cite{Nakamoto2008}.
+More generally it can be used to drive computation on a distributed virtual machine.
+In the Ethereum network, the virtual machine is used to implement smart contracts.
+The use here is much simpler: the simulation of a virtual world of generic cellular automata.
+
+It's certainly possible to run these automata on a centralized service, without using a blockchain.
+The main role that the blockchain plays is to establish a consensus for distributed simulation.
+
+The framework developed here is intended for recreational purposes: the coins mined by the system
 have no ostensible purpose other than regulating the rate at which participants can write to the simulated world.
 They might play the role of a score in a game world, measuring the contribution of the owner to keeping the world running.
-However, alternate scores (metrics) of success are possible in the world we describe, such as the popularity or success of programs that one has introduced into the world.
+However, alternate scores and success metrics are possible in the world we describe, such as the fitness of a user's programs.
 
@@
 I settled on the 8-bit MOS Technology 6502 processor for a CPU.
-The corresponding 16-bit memory map is shown in Table~\ref{tab:MemoryMap} and Figure~\ref{fig:UnitCube}.
+The corresponding 16-bit memory map is shown in Table~\ref{tab:MemoryMap}.
 
@@
   Memory map for the 6502 virtual machine of cell $(x,y,z)$,
-  using the 26-cell three-dimensional Moore neighborhood.
+  using the 26-cell three-dimensional Moore neighborhood (Figure~\ref{fig:UnitCube}).
   All address ranges not defined in this table are read-only and zero;

### 2018-07-03 paper.tex
period

@@
 If execution was terminated by timeout, then the rest of memory is also discarded,
-and the state of the CA is restored
+and the state of the CA is restored.
 Otherwise, if the program exited (by software interrupt or undefined opcode) before timing out,

### 2018-07-03 paper.tex
update
write

@@
 if their changes are to be committed to storage.
-As well as discouraging slow programs, this makes it easier to write programs that are atomic (all-or-nothing) in their effects.
+As well as discouraging crashes and other CPU hogs, this makes it easier to write programs that are atomic (all-or-nothing) in their effects.
 Programs can check the \code{TIME} timer at \hexrange{FFF6}{FFF7} to see how many cycles are available.
@@
 \subsection{Driving updates with the blockchain}
+\label{sec:ProofOfWork}
 
@@
 a {\bf write operation} which modifies the state of one or more cells of the cell block.
-This modification is considered to take place after the emulation of $Z$ update cycles
-which is part of the proof-of-work protocol.
+This modification is considered to take place after the emulation of $C$ update cycles
+which is part of the proof-of-work protocol (Section~\ref{sec:ProofOfWork}).
+
 The order in which such write operations are applied,
@@
 
-As well as writing the (volatile) state of the cell, you can write its 32K operating system
+As well as writing the (volatile) state of the cell, a write operation can overwrite its 32K operating system
 (mapped to address range \hexrange{8000}{FFEF} in the virtual machine; the last 16 bytes, corresponding to \hexrange{FFF0}{FFFF}, are ignored).

### 2018-07-03 paper.tex
write

@@
 
-As well as writing the (volatile) state of the cell, a write operation can overwrite its 32K operating system
+As well as writing the (volatile) state of the cell, a write operation can overwrite the cell's 32K operating system
 (mapped to address range \hexrange{8000}{FFEF} in the virtual machine; the last 16 bytes, corresponding to \hexrange{FFF0}{FFFF}, are ignored).

### 2018-07-03 paper.tex
adjacent cell blocks

@@
 
-The block may include sections of blockchain from {\bf adjacent cellular automata},
+The block may include sections of blockchain from {\bf adjacent cell blocks},
 defined to be cell blocks with co-ordinate ranges that

### 2018-07-04 paper.tex
ok

@@
 \usepackage{graphicx}
+\usepackage{amsfonts}
 
@@
 The cellular automata we develop as a motivating example are both universal and familiar, consisting of a three-dimensional grid of 6502 microprocessors broken into blocks.
-The blockchain framework incentivizes simulation of individual cell blocks, as well as data-sharing between blocks.
+The blockchain framework incentivizes simulation of individual chunks of cells, as well as traffic between chunks.
 
@@
 
-\subsection{Cell blocks}
+\subsection{Chunks of cells}
 
-Every cell of the cellular automata has a unique integer 3-space co-ordinate $(x,y,z)$.
+Every cell of the cellular automata has a unique integer 3-space co-ordinate $(x,y,z) \in \mathbb{Z}^3$.
 
-A single {\bf cell block} of the cellular automata is a $L \times L \times L$ array of cells,
-located at an offset $(\alpha_x,\alpha_y,\alpha_z)$ in integer 3-space.
-Cells in the cell block are indexed $(x,y,z$) where
-$\alpha_x \leq x < \alpha_x+L$,
-$\alpha_y \leq y < \alpha_y+L$,
-$\alpha_z \leq z < \alpha_z+L$.
+A single {\bf chunk} of the cellular automata is a $L \times L \times L$ cubic array of cells,
+located at an offset $L\alpha$,
+where $\alpha = (\alpha_x,\alpha_y,\alpha_z)$ is an integer 3-vector called the {\bf chunk location}.
+
+Cells in the chunk are indexed $(x,y,z$) where
+$L\alpha_x \leq x < (L+1)\alpha_x$,
+$L\alpha_y \leq y < (L+1)\alpha_y$,
+$L\alpha_z \leq z < (L+1)\alpha_z$.
 
@@
 All cells' bytes are initially zero, unless otherwise specified
-(including, implicitly, cells whose coordinates are outside the cell block; see also Section~\ref{sec:AdjacentCells}).
+(including, implicitly, cells whose coordinates are outside the chunk; see also Section~\ref{sec:AdjacentCells}).
 These bytes are the storage of the virtual machine.
@@
 
-\subsection{Updating the cell block}
-\label{sec:UpdatingCellBlock}
+\subsection{Updating the chunk}
+\label{sec:UpdatingChunk}
 
-An {\bf update} of the cell block consists of updating each cell in a pseudorandom order.
+An {\bf update} of the chunk consists of updating each cell in a pseudorandom order.
 This can be any order that visits $L^3$ uniformly-random selected cells in a well-defined pseudorandom order;
 a Mersenne Twister (MT19937),
-seeded on the tuple $(\alpha_x,\alpha_y,\alpha_z,L,U)$ where $U$ is the number of updates,
+seeded on the tuple $(\alpha,L,U)$ where $U$ is the number of updates,
 is a suitable random number generator (RNG) for this and other random numbers required during the update.
+
+Section~\ref{sec:AdjacentCells} describes a variant of this procedure whereby we construct a larger,
+virtual chunk (including adjacent chunks) in order to perform predictive simulation of the neighborhood.
 
@@
 that is transparently universal.
-The interface between the virtual machine and the cell block
+The interface between the virtual machine and the chunk storage
 needs to be such that the cell's ROM and the RAM, as well as the RAM of its neighbors,
@@
 
-The random variable \code{STOPTIME} is generated using the cell block RNG (Section~\ref{sec:UpdatingCellBlock})
+The random variable \code{STOPTIME} is generated using the chunk RNG (Section~\ref{sec:UpdatingChunk})
 and the following algorithm:
@@
 \item Let $z$ be the number of leading zero bits of $s$, so $0 \leq z \leq 16$
-\item Let $m = \mbox{max}(z,13) + 2$. The distribution of $m$ is now given by
+\item Let $m = \mbox{min}(z,13) + 2$. The distribution of $m$ is now given by
   \[
@@
 
-The machine architecture described so far is adequate if the operating system ROM is shared by every cell on the block.
-To allow O/S updates to be propagated across the block, we introduce two specific mechanisms.
+The machine architecture described so far is adequate if the operating system ROM is shared by every cell in the chunk.
+To allow O/S updates to be propagated across the chunk, we introduce two specific mechanisms.
 Guiding the design of this mechanism are the following two principles:
@@
 and then triggering a software interrupt, terminating execution.
-If the neighboring cell is outside the cell block (see Section~\ref{sec:AdjacentCells}),
+If the neighboring cell is outside the chunk (see Section~\ref{sec:AdjacentCells}),
 this updates the current cell (and its operating system) but has no effect on the neighboring cell.
 If the written byte is not a valid index byte, then there is no effect.
-A software interrupt is triggered, and program execution terminated, regardless of whether the written byte is a valid index pointing to a cell within the cell block.
+A software interrupt is triggered, and program execution terminated, regardless of whether the written byte is a valid index pointing to a cell within the chunk.
 
@@
 
-As mentioned above, the \code{SWAP} interface is one of two mechanisms by which operating system code is propagated across the cell block.
+As mentioned above, the \code{SWAP} interface is one of two mechanisms by which operating system code is propagated across the chunk.
 The second mechanism is built into the memory-mapped storage interface itself.
@@
 Reading a byte from \code{RAND} yields 8 random bits, newly sampled with each read (i.e. multiple reads yield different results).
-The random numbers are provided by the cell block's RNG (Section~\ref{sec:UpdatingCellBlock}).
+The random numbers are provided by the chunk's RNG (Section~\ref{sec:UpdatingChunk}).
 
@@
 
-\subsection{Adjacent cells}
+\subsection{Adjacent chunks}
 \label{sec:AdjacentCells}
 
-Consider those cell coordinates $(x,y,z)$ that are outside the cell block,
-e.g. $(\alpha_x-1,\alpha_y,\alpha_z)$ or  $(\alpha_x,\alpha_y,\alpha_z + L)$.
-Some of these, specifically those in the Moore neighborhood of cells that are in the cell block,
-are a relevant initial condition for the machine,
-since they map to bytes in its addressable memory.
+A chunk is {\bf adjacent} to the central chunk at $L\alpha$
+if it has a chunk location in the Moore neighborhood of $\alpha$
+(and therefore contains at least some cells in the neighborhood of cells in the central chunk).
 
-We refer to these cells as the cell block's {\bf adjacent cells}.
-Earlier, in Section~\ref{sec:Storage}, we specified that the state of these cells was implicitly initialized to zero.
-More generally, we may refer to the initial (and perpetual, throughout the execution of an update)
-state of these cells as the cell block's {\bf adjacent state}.
-This adjacent state is zero until otherwise specified, but may be affected
-by interactions with neighboring cell blocks in the infinite vector space,
-as for example by the blockchain of Section~\ref{sec:Blockchain}.
+When running several updates of the central chunk,
+given some knowledge (e.g. a recent snapshot) for the state of adjacent chunks,
+and assuming that communication with those adjacent chunks is imperfect (so that our snapshot may be out of date),
+we perform {\bf predictive simulation} of adjacent chunks.
+
+Specifically, when updating the central chunk, we also update our estimates of the adjacent chunks.
+To do this we construct a $(3L) \times (3L) \times (3L)$ {\bf virtual chunk},
+into which is mapped the current state of our chunk, and our current best estimates of the 26 adjacent chunks.
+
+We then update all 27 sub-chunks of the $3 \times 3 \times 3$ unit cube in the usual order ($z$-fastest, like the MSB in Table~\ref{tab:AddressBits}).
+Each sub-chunk's update is handled as though it were a single chunk,
+using the algorithm of Section~\ref{sec:UpdatingChunk},
+i.e. updating $L^3$ cells selected using a random number generator seeded with the tuple $(\alpha',L,U')$
+where $\alpha'$ is the adjacent chunk location and $U'$ is our estimate of the adjacent chunk update count
+(which is updated along with our estimate of the adjacent chunk states).
+
+Writes between adjacent chunks are not discarded, as they were previously
+(though writes outside of the $(3L) \times (3L) \times (3L)$ chunk will still be discarded).
+However, the central block's estimates of the state of its adjacent chunks
+may be overwritten in the future by news from those adjacent chunks,
+arriving in the blockchain in the form of adjacent segments (Section~\ref{sec:AdjacentSegments}).
+
+Predictive simulation facilitates traffic between chunks
+by allowing migration to be an active process, instead of a passive diffusion at the border.
 
@@
 
-The cell block has an associated blockchain, defined as in \cite{Nakamoto2008}.
+The chunk has an associated blockchain, defined as in \cite{Nakamoto2008}.
 Updates are incentivized by mining,
-and the reward for mining is adjusted to include redemption of coins mined by adjacent cell-block blockchains
-(this also updates the adjacent cell states accounted for by those cell blocks).
+and the reward for mining is adjusted to include redemption of coins mined by adjacent cell-chunk blockchains
+(this also updates the boundaries with those chunks).
 
@@
 
-A valid block must include the $256 \times L^3$ bytes describing the state of the cell block following $C$ updates
-from the state at the end of the previous block.
+A valid block must include a full description of the state of the $L^3$ cells in the central chunk,
+and the new best estimates for the $26L^3$ cells in adjacent chunks,
+following $C$ updates of the virtual chunk described in Section~\ref{sec:AdjacentCells}.
 $C$ is the number of cycles per cell expected in the blockchain update time
@@
 Let the definition of a block include the coordinates $(\alpha_x,\alpha_y,\alpha_z,L,U)$
-used to seed the cell block's random number generator and identify the cell block in space and time.
+used to seed the chunk's random number generator and identify the chunk in space and time,
+as well as the estimated update counts for adjacent chunks.
 
@@
 Any transaction in the blockchain may include, as additional data,
-a {\bf write operation} which modifies the state of one or more cells of the cell block.
+a {\bf write operation} which modifies the state of one or more cells of the chunk.
 This modification is considered to take place after the emulation of $C$ update cycles
@@
 \subsection{Adjacent segments}
+\label{sec:AdjacentSegments}
 
-The block may include sections of blockchain from {\bf adjacent cell blocks},
-defined to be cell blocks with co-ordinate ranges that
-overlap with the current cell block's adjacent cells,
-but no cells that overlap with
-cells in the co-ordinate range of this cell block (or any other cell block previously encountered in the blockchain,
-except the adjacent cell block itself).
-These blockchain segments from adjacent cell blocks are called {\bf adjacent segments}.
+The block may include sections of blockchain from adjacent chunks,
+referred to as {\bf adjacent segments}.
 
-Adjacent segments must be consistent with any previously-observed adjacent segments
-earlier on the current blockchain (that is, they must not involve conflicting cell block,
-and they must otherwise be consistent continuations of the adjacent cell block's blockchain).
+Adjacent segments must be consistent continuations of any previously-observed adjacent segments
+earlier on the central blockchain.
 They may, themselves, recursively refer to parts of the current blockchain as their own adjacent segments,
@@
 
-When a block includes an adjacent segment, the cell block's adjacent cell state is updated to match the adjacent segment.
-Regular mutual inclusion of adjacent segments from adjacent cell blocks' blockchains
-corresponds to keeping a bidirectional read-only interface open between those cell blocks.
+When a block includes an adjacent segment, the chunk's estimate for that adjacent chunk's state is updated to match the information in the adjacent segment.
+Regular mutual inclusion of adjacent segments from adjacent chunks' blockchains
+corresponds to keeping a bidirectional read-only interface open between those chunks.
 
-To incentivize this,
+To incentivize this bidirectional traffic,
 inclusion of an adjacent segment in a block for the first time
@@
 where $C$ is the number of {\bf redeemable coins} mined or accrued (via more distant, proximal adjacencies)
-by cells in the adjacent cell block,
-and $R$ is the {\bf redemption rate}, the number of redeemable coins in adjacent cell blocks
-required to accrue a coin in the current cell block.
+by cells in the adjacent chunk,
+and $R$ is the {\bf redemption rate}, the number of redeemable coins in adjacent chunks
+required to accrue a coin in the current chunk.
 It is of course necessary to have $R>2$ to prevent an explosion of redeemable coins in 3D space;
@@
 
-Setting the parameters $L$ (cell block size) and $C$ (updates per block)
+Setting the parameters $L$ (chunk size) and $C$ (updates per block)
 is a little arbitrary, but can be guided by target memory footprint and clock rate.
-For memory, the cell block state requires $256 \times L^3$ bytes.
+For memory, the chunk state requires $256 \times L^3$ bytes.
 For clock rate, if we ran a 6502 CPU at 1Mhz for 10 minutes, with a mean of $44.5$ cycles per cellular update,
-we would update the whole cell block $10^6 \times T / (44.5 \times L^3)$ times
+we would update the whole chunk $10^6 \times T / (44.5 \times L^3)$ times
 where $T$ is the time in seconds between blocks ($T=36,000$ for bitcoin).
@@
 each cell is updated every few minutes;
-each block takes at most 64Mb to describe the cell block state;
+each block takes at most 64Mb to describe the chunk state;
 and the computational burden is roughly comparable to a 6502 CPU.
@@
   writing to these addresses has no effect.
-  Pages mapping to cells outside the cell block (e.g. $x<\alpha_x$ or $x \geq \alpha_x+L$, or similarly for $y,z$)
+  Pages mapping to cells outside the chunk (e.g. $x<L\alpha_x$ or $x \geq (L+1)\alpha_x$, or similarly for $y,z$)
   are read-only and zero (unless otherwise specified, as in Section~\ref{sec:AdjacentCells}).

### 2018-07-06 paper.tex
labels

@@
 \newcommand\code[1]{{\tt #1}}
-\newcommand\hex[1]{{\tt #1}}
+\newcommand\boldcode[1]{\code{#1}}
+\newcommand\hex[1]{{\tt 0x#1}}
 \newcommand\hexrange[2]{\hex{#1}{\tt -}\hex{#2}}
@@
 \newcommand\memtable[1]{
-\begin{tabular}{lll}
+\begin{tabular}{rlll}
   \hline
-  From & To & Contents \\
+  Label & From & To & Contents \\
   \hline
@@
 
-\newcommand\memrow[4]{
-    \hex{{#1}00} & \hex{{#1}FF} & State of cell $(x#2,y#3,z#4)$ \\
+\newcommand\memrow[5]{
+    \boldcode{#5} & \hex{{#1}00} & \hex{{#1}FF} & State of cell $(x#2,y#3,z#4)$ \\
 }
@@
 % The \memxyz macro gives a densely-connected (von Neumann) neighborhood
-\newcommand\memz[5]{
-  \memrow{#1}{#4}{#5}{-1}
-  \memrow{#2}{#4}{#5}{}
-  \memrow{#3}{#4}{#5}{+1}
+\newcommand\memz[6]{
+  \memrow{#1}{#4}{#5}{-1}{D#6}
+  \memrow{#2}{#4}{#5}{}{#6}
+  \memrow{#3}{#4}{#5}{+1}{U#6}
 }
 
-\newcommand\memyz[2]{
-  \memz{{#1}0}{{#1}1}{{#1}2}{#2}{-1}
-  \memz{{#1}4}{{#1}5}{{#1}6}{#2}{}
-  \memz{{#1}8}{{#1}9}{{#1}A}{#2}{+1}
+\newcommand\memyz[3]{
+  \memz{{#1}0}{{#1}1}{{#1}2}{#2}{-1}{S#3}
+  \memz{{#1}4}{{#1}5}{{#1}6}{#2}{}{#3}
+  \memz{{#1}8}{{#1}9}{{#1}A}{#2}{+1}{N#3}
 }
@@
 \newcommand\memxyz{
-  \memyz{4}{-1}
+  \memyz{4}{-1}{W}
 
   %  \memyz{5}{} includes zero page-mapped cell (x,y,z), which we want to skip, so spell it out without that
-  \memz{50}{51}{52}{}{-1}
-  \memrow{54}{}{}{-1}
-  \memrow{56}{}{}{+1}
-  \memz{58}{59}{5A}{}{+1}
+  \memz{50}{51}{52}{}{-1}{S}
+  \memrow{54}{}{}{-1}{W}
+  \memrow{56}{}{}{+1}{E}
+  \memz{58}{59}{5A}{}{+1}{N}
 
-  \memyz{6}{+1}
+  \memyz{6}{+1}{E}
 }
 
-% The following table is a sparse neighborhood
+% The following table is a sparse neighborhood (von Neumann, 6 neighbors/cell)
 \newcommand\vonneumannmap{\memtable{
-  \memrow{45}{-1}{}{}
-  \memrow{51}{}{-1}{}
-  \memrow{54}{}{}{-1}
-  \memrow{56}{}{}{+1}
-  \memrow{59}{}{+1}{}
-  \memrow{65}{+1}{}{}
+  \memrow{45}{-1}{}{}{W}
+  \memrow{51}{}{-1}{}{S}
+  \memrow{54}{}{}{-1}{D}
+  \memrow{56}{}{}{+1}{U}
+  \memrow{59}{}{+1}{}{N}
+  \memrow{65}{+1}{}{}{E}
 }}
 
-% The following neighborhood is dense
+% The following neighborhood is dense (Moore, 26 neighbors/cell)
 \newcommand\mooremap{\memtable{\memxyz}}
+
+% Captions
+\newcommand\labeq[2]{\boldcode{#1}\code{=}\hex{#2}}
+\newcommand\asslabels{
+  The labels (\code{N}, \code{S}, \code{E}, \code{W}, \code{U}, \code{D} \ldots)
+  are pre-defined in the assembler to the respective base addresses (\labeq{N}{5900}, \labeq{S}{5100}, and so on).
+  The corresponding labels with \code{PAGE} appended
+  are predefined to be the respective MSBs of those base addresses (\labeq{NPAGE}{59}, \labeq{SPAGE}{51}, and so on).
+  The assembler also pre-defines the symbols
+  \labeq{XDIR}{1000}, \labeq{YDIR}{0400}, \labeq{ZDIR}{0100};
+  \labeq{XPAGE}{10}, \labeq{YPAGE}{04}, \labeq{ZPAGE}{01};
+  \labeq{ORIG}{5500}, \labeq{ORIGPAGE}{55};
+  \labeq{BASE}{4000}, \labeq{BASEPAGE}{40}.
+}
 
@@
   \hex{FFF6} & \hex{FFF7} & Read-only & \code{TIME}: number of cycles left until termination \\
-  \hex{FFF8} & \hex{FFF8} & Write-only & \code{SWAP}: swap storage page with neighbor \\
+  \hex{FFF8} & \hex{FFF8} & Write-only & \code{SWAP}: write to swap cell with neighbor \\
   \hex{FFF9} & \hex{FFF9} & Read-only & \code{RAND}: random byte generator \\
@@
   Writing to these addresses has no effect.
+  \asslabels
 }
@@
   This memory map may be considered a subset of the memory map in Table~\ref{tab:MooreNeighborhood}.
+  \asslabels
 }
@@
 \caption{
-  The unit cube of the Moore neighborhood, and its relationship to the memory map of Table~\ref{tab:MooreNeighborhood}.
+  The memory-mapped neighborhood of a cell (see Table~\ref{tab:MemoryMap} and Table~\ref{tab:MooreNeighborhood}).
   The central cell is at $(1,1,1)$, and is mapped to zero page.
@@
   the first six bits of the address MSB corresponding to the $x$, $y$ and $z$ co-ordinates within this cell (see Table~\ref{tab:AddressBits}).
+  \asslabels
   \label{fig:UnitCube}

### 2018-07-06 paper.tex
updated

@@
 A single {\bf chunk} of the cellular automata is a $L \times L \times L$ cubic array of cells,
-located at an offset $L\alpha$,
+located at a base offset $L\alpha$,
 where $\alpha = (\alpha_x,\alpha_y,\alpha_z)$ is an integer 3-vector called the {\bf chunk location}.
@@
 Each cell has 256 bytes (one page) of updatable state (RAM).
-All cells' bytes are initially zero, unless otherwise specified
-(including, implicitly, cells whose coordinates are outside the chunk; see also Section~\ref{sec:AdjacentCells}).
+All cells' bytes are initially zero
+(including, implicitly, cells whose coordinates are outside the chunk; see also Section~\ref{sec:NeighborChunks}).
 These bytes are the storage of the virtual machine.
@@
 
-An {\bf update} of the chunk consists of updating each cell in a pseudorandom order.
-This can be any order that visits $L^3$ uniformly-random selected cells in a well-defined pseudorandom order;
-a Mersenne Twister (MT19937),
+An {\bf update} of the chunk consists of updating $L^3$ uniformly-random selected cells in a well-defined pseudorandom order.
+A Mersenne Twister (MT19937),
 seeded on the tuple $(\alpha,L,U)$ where $U$ is the number of updates,
-is a suitable random number generator (RNG) for this and other random numbers required during the update.
+is a suitable random number generator (RNG) for this
+and for generating the random cycle limit \code{STOPTIME} (Section~\ref{sec:Timeouts}),
+which also must be generated for every updated cell.
 
-Section~\ref{sec:AdjacentCells} describes a variant of this procedure whereby we construct a larger,
-virtual chunk (including adjacent chunks) in order to perform predictive simulation of the neighborhood.
+Section~\ref{sec:NeighborChunks} describes a variant of this procedure whereby we construct a larger,
+virtual chunk (including neighboring chunks) in order to perform predictive simulation of the neighborhood.
 
@@
 and then triggering a software interrupt, terminating execution.
-If the neighboring cell is outside the chunk (see Section~\ref{sec:AdjacentCells}),
-this updates the current cell (and its operating system) but has no effect on the neighboring cell.
+If the neighboring cell is outside the chunk,
+this updates the current cell (and its operating system) but has no effect on the neighboring cell
+(although see Section~\ref{sec:NeighborChunks}).
 If the written byte is not a valid index byte, then there is no effect.
@@
 Reading a byte from \code{RAND} yields 8 random bits, newly sampled with each read (i.e. multiple reads yield different results).
-The random numbers are provided by the chunk's RNG (Section~\ref{sec:UpdatingChunk}).
+
+The random numbers are provided by an RNG seeded on the cell and chunk co-ordinates and the chunk update count
+(NB this is a per-cell RNG, distinct from the per-chunk RNG of Section~\ref{sec:UpdatingChunk} that selects cells and sets \code{STOPTIME}).
 
@@
 
-\subsection{Adjacent chunks}
-\label{sec:AdjacentCells}
+\subsection{Neighboring chunks}
+\label{sec:NeighborChunks}
 
-A chunk is {\bf adjacent} to the central chunk at $L\alpha$
-if it has a chunk location in the Moore neighborhood of $\alpha$
-(and therefore contains at least some cells in the neighborhood of cells in the central chunk).
+Two $L$-sized chunks are neighbors if their chunk locations are neighbors;
+that is, they have base offsets $L\alpha_1$ and $L\alpha_2$,
+and $\alpha_1$ is in the Moore neighborhood of $\alpha_2$.
+This is equivalent to the statement that (at least) one cell in the first chunk has (at least) one cell in the second chunk in its Moore neighborhood, and vice versa.
 
 When running several updates of the central chunk,
-given some knowledge (e.g. a recent snapshot) for the state of adjacent chunks,
-and assuming that communication with those adjacent chunks is imperfect (so that our snapshot may be out of date),
-we perform {\bf predictive simulation} of adjacent chunks.
+given some knowledge (e.g. a recent snapshot) for the state of neighboring chunks,
+and assuming that communication with those neighboring chunks is imperfect (so that our snapshot may be out of date),
+we perform {\bf predictive simulation} of neighboring chunks.
 
-Specifically, when updating the central chunk, we also update our estimates of the adjacent chunks.
-To do this we construct a $(3L) \times (3L) \times (3L)$ {\bf virtual chunk},
-into which is mapped the current state of our chunk, and our current best estimates of the 26 adjacent chunks.
+Specifically, when updating the central chunk, we also update our estimates of the neighboring chunks.
+To do this we construct a $(3L) \times (3L) \times (3L)$ virtual {\bf super-chunk},
+into which is mapped the current state of our chunk, as well as our current best estimates of the 26 neighboring chunks.
 
@@
 i.e. updating $L^3$ cells selected using a random number generator seeded with the tuple $(\alpha',L,U')$
-where $\alpha'$ is the adjacent chunk location and $U'$ is our estimate of the adjacent chunk update count
-(which is updated along with our estimate of the adjacent chunk states).
+where $\alpha'$ is the neighboring chunk location and $U'$ is our estimate of the neighboring chunk update count
+(which is updated along with our estimate of the neighboring chunk states).
 
-Writes between adjacent chunks are not discarded, as they were previously
+Writes between neighboring chunks are not discarded, as they were previously
 (though writes outside of the $(3L) \times (3L) \times (3L)$ chunk will still be discarded).
-However, the central block's estimates of the state of its adjacent chunks
-may be overwritten in the future by news from those adjacent chunks,
-arriving in the blockchain in the form of adjacent segments (Section~\ref{sec:AdjacentSegments}).
+However, the central block's estimates of the state of its neighboring chunks
+may be overwritten in the future by news from those neighboring chunks,
+arriving in the blockchain in the form of neighboring segments (Section~\ref{sec:NeighboringSegments}).
 
@@
 Updates are incentivized by mining,
-and the reward for mining is adjusted to include redemption of coins mined by adjacent cell-chunk blockchains
+and the reward for mining is adjusted to include redemption of coins mined by neighboring cell-chunk blockchains
 (this also updates the boundaries with those chunks).
@@
 A valid block must include a full description of the state of the $L^3$ cells in the central chunk,
-and the new best estimates for the $26L^3$ cells in adjacent chunks,
-following $C$ updates of the virtual chunk described in Section~\ref{sec:AdjacentCells}.
+and the new best estimates for the $26L^3$ cells in neighboring chunks,
+following $C$ updates of the super-chunk described in Section~\ref{sec:NeighborChunks}.
 $C$ is the number of cycles per cell expected in the blockchain update time
@@
 used to seed the chunk's random number generator and identify the chunk in space and time,
-as well as the estimated update counts for adjacent chunks.
+as well as the estimated update counts for neighboring chunks.
 
@@
 
-\subsection{Adjacent segments}
-\label{sec:AdjacentSegments}
+\subsection{Neighboring segments}
+\label{sec:NeighboringSegments}
 
-The block may include sections of blockchain from adjacent chunks,
-referred to as {\bf adjacent segments}.
+The block may include sections of blockchain from neighboring chunks,
+referred to as {\bf neighboring segments}.
 
-Adjacent segments must be consistent continuations of any previously-observed adjacent segments
+Neighboring segments must be consistent continuations of any previously-observed neighboring segments
 earlier on the central blockchain.
-They may, themselves, recursively refer to parts of the current blockchain as their own adjacent segments,
+They may, themselves, recursively refer to parts of the current blockchain as their own neighboring segments,
 summarized by their hashes for compactness.
 
-When a block includes an adjacent segment, the chunk's estimate for that adjacent chunk's state is updated to match the information in the adjacent segment.
-Regular mutual inclusion of adjacent segments from adjacent chunks' blockchains
+When a block includes an neighboring segment, the chunk's estimate for that neighboring chunk's state is updated to match the information in the neighboring segment.
+Regular mutual inclusion of neighboring segments from neighboring chunks' blockchains
 corresponds to keeping a bidirectional read-only interface open between those chunks.
@@
 To incentivize this bidirectional traffic,
-inclusion of an adjacent segment in a block for the first time
+inclusion of an neighboring segment in a block for the first time
 accrues an additional benefit of $C/R$ coins to the miner of the block,
-where $C$ is the number of {\bf redeemable coins} mined or accrued (via more distant, proximal adjacencies)
-by cells in the adjacent chunk,
-and $R$ is the {\bf redemption rate}, the number of redeemable coins in adjacent chunks
+where $C$ is the number of {\bf redeemable coins} mined or accrued (via transitive neighbor relationships)
+by cells in the neighboring chunk,
+and $R$ is the {\bf redemption rate}, the number of redeemable coins in neighboring chunks
 required to accrue a coin in the current chunk.
-It is of course necessary to have $R>2$ to prevent an explosion of redeemable coins in 3D space;
-another notable value is $R=26$, which (in the densely-connected Moore neighborhood)
+It is of course necessary to have $R>2$ to prevent an explosion of redeemable value in 3D space;
+a conservative value is $R=26$, which (in the densely-connected Moore neighborhood)
 ensures that the number of coins redeemed is less than the number of coins mined locally
@@
   Pages mapping to cells outside the chunk (e.g. $x<L\alpha_x$ or $x \geq (L+1)\alpha_x$, or similarly for $y,z$)
-  are read-only and zero (unless otherwise specified, as in Section~\ref{sec:AdjacentCells}).
+  are read-only and zero (unless otherwise specified, as in Section~\ref{sec:NeighborChunks}).
   Writing to these addresses has no effect.

### 2018-07-07 cpu.js
imported CPU emulator of Torlus/6502.js

@@ -1,0 +1,669 @@
+// based on https://github.com/Torlus/6502.js/blob/master/src/cpu.js
+// crudely hacked back to ES5
+
+var CPU6502op = require ('./opcodes.js')
+
+
+/**
+ * A 6502 processor emulator
+ */
+class CPU6502{
+    constructor(){
+        this.PC = 0; // Program counter
+
+	    this.A = 0; this.X = 0; this.Y = 0; this.S = 0; // Registers
+	    this.N = 0; this.Z = 1; this.C = 0; this.V = 0; // ALU flags
+	    this.I = 0; this.D = 0; // Other flags
+
+	    this.irq = 0; this.nmi = 0; // IRQ lines
+
+	    this.tmp = 0; this.addr = 0; // Temporary registers
+	    this.opcode = 0; // Current opcode
+	    this.cycles = 0; // Cycles counter
+    }
+    
+    ////////////////////////////////////////////////////////////////////////////////
+    // CPU control
+    ////////////////////////////////////////////////////////////////////////////////
+    
+    /**
+     * Reset the processor
+     */
+    reset() {
+
+	    this.A = 0; this.X = 0; this.Y = 0; this.S = 0;
+	    this.N = 0; this.Z = 1; this.C = 0; this.V = 0;
+	    this.I = 0; this.D = 0;
+
+	    this.PC = (this.read(0xFFFD) << 8) | this.read(0xFFFC);
+    }
+    
+    /**
+     * Execute a single opcode
+     */
+    step() {
+	    this.opcode = this.read( this.PC++ );
+	    CPU6502op[ this.opcode ]( this );
+    }
+    
+    /**
+     * Log the current cycle count and all registers to console.log
+     */
+    log(){
+	    var msg = "nPC=" + this.PC.toString(16);
+	    msg += " cyc=" + this.cycles;
+	    msg += " [" + this.opcode.toString(16) + "] ";
+	    msg += ( this.C ? "C" : "-");
+	    msg += ( this.N ? "N" : "-");
+	    msg += ( this.Z ? "Z" : "-");
+	    msg += ( this.V ? "V" : "-");
+	    msg += ( this.D ? "D" : "-");
+	    msg += ( this.I ? "I" : "-");
[…]

### 2018-07-07 opcodes.js

@@ -1,0 +1,282 @@
+// based on https://github.com/Torlus/6502.js/blob/master/src/opcodes.js
+// bludgeoned back to ES5, sorry
+
+////////////////////////////////////////////////////////////////////////////////
+// Opcode table
+////////////////////////////////////////////////////////////////////////////////
+
+var CPU6502op = new Array();
+
+/*  BRK     */ CPU6502op[0x00] = function(m) { m.imp(); m.brk(); };
+/*  ORA izx */ CPU6502op[0x01] = function(m) { m.izx(); m.ora(); };
+/* *KIL     */ CPU6502op[0x02] = function(m) { m.imp(); m.kil(); };
+/* *SLO izx */ CPU6502op[0x03] = function(m) { m.izx(); m.slo(); m.rmw(); };
+/* *NOP zp  */ CPU6502op[0x04] = function(m) { m.zp(); m.nop(); };
+/*  ORA zp  */ CPU6502op[0x05] = function(m) { m.zp(); m.ora(); };
+/*  ASL zp  */ CPU6502op[0x06] = function(m) { m.zp(); m.asl(); m.rmw(); };
+/* *SLO zp  */ CPU6502op[0x07] = function(m) { m.zp(); m.slo(); m.rmw(); };
+/*  PHP     */ CPU6502op[0x08] = function(m) { m.imp(); m.php(); };
+/*  ORA imm */ CPU6502op[0x09] = function(m) { m.imm(); m.ora(); };
+/*  ASL     */ CPU6502op[0x0A] = function(m) { m.imp(); m.asla(); };
+/* *ANC imm */ CPU6502op[0x0B] = function(m) { m.imm(); m.anc(); };
+/* *NOP abs */ CPU6502op[0x0C] = function(m) { m.abs(); m.nop(); };
+/*  ORA abs */ CPU6502op[0x0D] = function(m) { m.abs(); m.ora(); };
+/*  ASL abs */ CPU6502op[0x0E] = function(m) { m.abs(); m.asl(); m.rmw(); };
+/* *SLO abs */ CPU6502op[0x0F] = function(m) { m.abs(); m.slo(); m.rmw(); };
+
+/*  BPL rel */ CPU6502op[0x10] = function(m) { m.rel(); m.bpl(); };
[…]

### 2018-07-07 asm.js
assembler

@@ -1,0 +1,601 @@
+// 6502 assembler
+// n. landsteiner, mass:werk / electronic tradion 2005; e-tradion.net
+
+// lookup tables
+
+var hextab= ['0','1','2','3','4','5','6','7','8','9','A','B','C','D','E','F'];
+var instrtab = {
+	ADC:[  -1,  -1,0x69,0x6d,0x7d,0x79,0x65,0x75,  -1,  -1,0x61,0x71,  -1],
+	AND:[  -1,  -1,0x29,0x2d,0x3d,0x39,0x25,0x35,  -1,  -1,0x21,0x31,  -1],
+	ASL:[  -1,0x0a,  -1,0x0e,0x1e,  -1,0x06,0x16,  -1,  -1,  -1,  -1,  -1],
+	BCC:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0x90],
+	BCS:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0xb0],
+	BEQ:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0xf0],
+	BIT:[  -1,  -1,  -1,0x2c,  -1,  -1,0x24,  -1,  -1,  -1,  -1,  -1,  -1],
+	BMI:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0x30],
+	BNE:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0xd0],
+	BPL:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0x10],
+	BRK:[0x00,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1],
+	BVC:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0x50],
+	BVS:[  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,  -1,0x70],
[…]

### 2018-07-07 src/chunk.js
src/chunk.js

@@ -1,0 +1,13 @@
+var CPU6502 = require('./cpu')
+
+class Chunk {
+  constructor (config) {
+    config = config || {}
+    this.xSize = config.xSize || config.size || 1
+    this.ySize = config.ySize || config.size || 1
+    this.zSize = config.zSize || config.size || 1
+    // More to go here
+  }
+}
+
+module.exports = Chunk;

### 2018-07-07 src/chunk.js
src/mersenne-twister.js

@@ -6,7 +6,13 @@ class Chunk {
     this.xSize = config.xSize || config.size || 1
     this.ySize = config.ySize || config.size || 1
     this.zSize = config.zSize || config.size || 1
-    // More to go here
+    this.state = new Array(this.xSize).fill(0).map (function() {
+      return new Array(this.ySize).fill(0).map (function() {
+        return new Array(this.zSize).fill(0).map (function() {
+          return new Uint8Array (256)
+        })
+      })
+    })
   }
 }
 

### 2018-07-07 src/chunk.js
src/chunk.js

@@ -1,18 +1,168 @@
 var CPU6502 = require('./cpu')
+var MersenneTwister = require('./mersenne-twister')
+
+var defaultRngSeed = 0x12345678
 
 class Chunk {
   constructor (config) {
+    var chunk = this
     config = config || {}
-    this.xSize = config.xSize || config.size || 1
-    this.ySize = config.ySize || config.size || 1
-    this.zSize = config.zSize || config.size || 1
-    this.state = new Array(this.xSize).fill(0).map (function() {
-      return new Array(this.ySize).fill(0).map (function() {
-        return new Array(this.zSize).fill(0).map (function() {
+    function makeSize (sz) {
+      sz = sz || config.size || 1
+      if (Math.pow (2, Math.floor (Math.log(sz) / Math.log(2))) !== sz)
+        throw new Error ("Size " + sz + " is not a power of 2")
+      return sz
+    }
+    this.xSize = makeSize (config.xSize)
+    this.ySize = makeSize (config.ySize)
+    this.zSize = makeSize (config.zSize)
+    this.cell = new Array(chunk.xSize).fill(0).map (function() {
+      return new Array(chunk.ySize).fill(0).map (function() {
+        return new Array(chunk.zSize).fill(0).map (function() {
           return new Uint8Array (256)
         })
       })
     })
+    this.rng = new MersenneTwister (typeof(config.seed) === 'undefined' ? defaultRngSeed : config.seed)
+    this.pageDelta = [[-1,-1,-1], [-1,-1,0], [-1,-1,+1],
+                      [-1, 0,-1], [-1, 0,0], [-1, 0,+1],
+                      [-1,+1,-1], [-1,+1,0], [-1,+1,+1],
+                       
+                      [ 0,-1,-1], [ 0,-1,0], [ 0,-1,+1],
+                      [ 0, 0,-1], /* zero */ [ 0, 0,+1],
+                      [ 0,+1,-1], [ 0,+1,0], [ 0,+1,+1],
+
+                      [+1,-1,-1], [+1,-1,0], [+1,-1,+1],
+                      [+1, 0,-1], [+1, 0,0], [+1, 0,+1],
+                      [+1,+1,-1], [+1,+1,0], [+1,+1,+1]]
+  }
+
+  numberOfCells() {
+    return this.xSize * this.ySize * this.zSize
+  }
+  
+  cellPage (x, y, z) {
[…]
